---
title: "ps-processing"
author: "Maggie Sogin"
date: "2026-06-02"
output:
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
    toc_float: true
---

# Set up working space
read in packages
```{r}
library(tidyverse)
library(phyloseq);packageVersion('phyloseq')
library(ggplot2)
library(decontam);packageVersion('decontam')
library(microbiome);packageVersion('microbiome')
library(microViz);packageVersion('microViz')
```
set up plotting conditions
```{r}
#set up plotting conditions
trt<-c("FASW","ABS","labrenzia","ruegeria","vibrio","ISOABC")
myPal<-c('grey60','black','#A6CEE3','#1F78B4','#B2DF8A','violet')
names(myPal) = trt

shpPal<-c(15,16,17,18,4,1)
dpis<-c("21","14","4","1","0","-1")
names(shpPal) = dpis

dpiPal<-c('black','gray30','seashell','seashell2','seashell3','seashell4')
names(dpiPal) <- c('-1','0','1','4','14','21')

theme_new<- theme_bw() + theme(panel.grid.major = element_line(color='gray90'), panel.grid.minor = element_line(color='gray90'), text= element_text(size = 15), axis.text.x = element_text(angle = 65,hjust = 1), legend.text = element_text(size = 20), axis.title = element_text(size = 15), title = element_text(size = 20))
```

# Read in and manipulate data

Notes: ps object was generated using dada2.mock.r and dada2.samples.r pipeline with the
following notes: - error models was learned on mock samples - error
model was applied to all experimental samples to learn error - sequence
stats are reported in stats.csv

chimera detection notes: When including the mock community samples which are rich in chimeras, we detected 1043 chimeric sequences out of 2081 ASVs. When we run the pipeline without the mock communities, only 53 ASVs out of 988 AVS are determine to be chimeras

## Read in data

```{r}
ps<-readRDS('../../data/16S/ps.withTree.rds')
df.meta<-read.csv('../../data/16S/metadata.csv')
ps
head(df.meta); dim(df.meta)
```
## Sample data frame

```{r}
#sample names
tmp<-data.frame(sample_data(ps))
tmp$sample.names<-str_replace(tmp$sample.names,'_pacbio.fastq.gz','')
tmp$names.rows<-rownames(tmp)

#join two datasets
tmp.full<-left_join(tmp,df.meta, by='sample.names')
head(tmp.full); tail(tmp.full)
```

```{r}
#rename rows
rownames(tmp.full)<-tmp.full$names.rows

#rename as metadata.df and remove names.rows column
metadata.df<-tmp.full %>% dplyr::select(sample.names,treatment,timepoint,DNA_conc_ng_ul)

head(metadata.df); tail(metadata.df); dim(metadata.df)

#set factor levels for treatment and dpi
metadata.df<-metadata.df %>% 
  mutate(dpi=case_when(timepoint=='post_priming' ~ -1,
                      timepoint=='post_ABS' ~0,
                      timepoint=='1dpi' ~ 1,
                      timepoint=='4dpi' ~ 4,
                      timepoint=='14dpi' ~ 14,
                      timepoint=='21dpi' ~ 21 )) %>% 
  mutate(dpi=factor(dpi,levels=c(21,14,4,1,0,-1, NA),ordered = T)) %>% 
  arrange(dpi) %>% 
  mutate(sample.names2=factor(sample.names, levels=rev(unique(sample.names)))) %>% 
  mutate(sample.names3=paste(sample.names2, dpi)) %>% 
  mutate(treatment=factor(x=treatment, levels=c('FASW','ABS','ISOABC','labrenzia','ruegeria','vibrio','ext_blk', 'mock_extracted','mock_dna')))

#add to phyloseq object
sample_data(ps)<-metadata.df
ps
```
# Preprocess data
## Remove blank wells
```{r}
ps<-ps %>% subset_samples(!is.na(treatment))
```

## Remove chloroplast and mitocondria reads
```{r}
ps_clean <- ps %>%
  subset_taxa(
    Order != "Chloroplast" &
    Family != "Mitochondria" &
    Class != "Chloroplast" # Sometimes Chloroplast is in Class
  )
ps
ps_clean # went from 1038 ASVs to 923 AVS
```
## Run decontam pipeline
use frequency method with threshold set at 0.1

```{r}
df <- as.data.frame(sample_data(ps_clean)) # Put sample_data into a ggplot-friendly data.frame
df$sample_or_control<-'sample'
df[df$treatment=='ext_blk','sample_or_control']<-'control'
df$LibrarySize <- sample_sums(ps_clean)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=sample_or_control)) + geom_point()
```

Identify contaminants using Prevalence method, with threshold set at 0.5
```{r}
sample_data(ps_clean)<-df
sample_data(ps_clean)$is.neg <- sample_data(ps_clean)$sample_or_control == "control"
contamdf<- isContaminant(ps_clean, 
                         method="frequency", 
                         neg="is.neg", 
                         threshold=0.1,
                         conc="DNA_conc_ng_ul",
                         )
table(contamdf$contaminant)
(ct<-rownames(contamdf)[which(contamdf$contaminant)])
```

Need to investigate taxa deemed contaminants
```{r}
tax_table(subset_taxa(ps_clean,rownames(tax_table(ps_clean)) %in% ct))
```
remove contaminants 
```{r}
ps_clean.2<-subset_taxa(ps_clean,!rownames(tax_table(ps_clean)) %in% ct)
```

## Prune samples
Investigate the number of reads per sample across treatments

```{r}
readsums.df <- data.frame(nreads = sort(sample_sums(ps_clean.2), TRUE), sorted = 1:nsamples(ps_clean.2), type = "sample")
readsums.df <- readsums.df %>% mutate(sample.names=str_replace(rownames(readsums.df),'_pacbio.fastq.gz',''))
readsums.df <- full_join(readsums.df,df.meta, by='sample.names')

#plot across treatments
ggplot(readsums.df, aes(x=treatment, y=nreads)) + geom_point() + ylim(0,10000) #look at samples with fewer then 10K reads
```
only plotting low read number samples 

remove samples that have less then 2500 reads per sample
```{r}
ps_clean.2.5Kreads<-prune_samples(sample_sums(ps_clean.2) >=2500, ps_clean.2)
ps_clean.2.5Kreads #goes from 128 to 105 samples

#get treatment size
samps.sizes<-data.frame(sample_data(ps_clean.2.5Kreads)) %>% group_by(treatment, timepoint) %>% summarise(count=length(sample.names))
samps.sizes %>% arrange(count)
```
```{r}
summarize_phyloseq(ps_clean.2.5Kreads)
```

## Fix taxa names and validate ps object

```{r}
ps.all<-ps_clean.2.5Kreads  %>% tax_fix() %>% phyloseq_validate(remove_undetected = T)
ps.all
```

## Subset out experimental samples
```{r}
ps.exp<-prune_samples(!is.na(sample_data(ps_clean.2.5Kreads)$timepoint),ps_clean.2.5Kreads)
ps.exp<-prune_taxa(taxa_sums(ps.exp)>0, ps.exp) #remove ASVs with 0 counts
ps.exp 
samps.sizes<-data.frame(sample_data(ps.exp)) %>% group_by(treatment, timepoint) %>% summarise(count=length(sample.names))
samps.sizes 
```


## Save phyloseq object
```{r}
saveRDS(ps.exp, file='../data/16S/ps.exp.samples.rds')
```

# Mock data analysis

## subset out mock samples
```{r}
ps.mock<-prune_samples(sample_data(ps_clean.2.5Kreads)$treatment %in% c('mock_extracted','mock_dna'),ps_clean.2.5Kreads)
ps.mock<-prune_taxa(taxa_sums(ps.mock)>0, ps.mock) #remove ASVs with 0 counts
ps.mock 
samps.sizes<-data.frame(sample_data(ps.mock)) %>% group_by(treatment, timepoint) %>% summarise(count=length(sample.names))
samps.sizes 
```
## barplot 

```{r}
ps.mock %>% 
  tax_fix() %>% 
  tax_filter(min_prevalence = 0.5,min_total_abundance = 0.01, tax_level = "Species")  %>% 
  comp_barplot(tax_level = 'Species', n_taxa = 15,facet_by = "treatment") +
  theme(axis.text.x = element_blank())
```
Expected taxa for mock: 
Bacillus subtilis  ✅
Listeria monocytogenes  ✅
staphylococcus aureus ✔️
Lactobacillus fermentum (aka Limosilactobacillus) ✅
Salmonella enterica  ✅
E. coli ✅
Enterococcus faecalis ✅
Pseudomonas aeruginosa ✅

There are also some other things but generally this looks good



