---
title: "statistical_analysis_16SrRNA"
author: "Maggie Sogin"
date: "June 9 2026"
output:
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
    toc_float: true
---

# Set up working space
### Read in packages
```{r}
#load in packages and print versions 
packages <- c("tidyverse", "ggplot2", "vegan", "phyloseq", "microbiome",
               "ranacapa", "microViz", "patchwork", "pairwiseAdonis", "pheatmap","ALDEx2","emmeans")
invisible(lapply(packages, function(pkg) {
  library(pkg, character.only = TRUE)
  cat(pkg, ":", as.character(packageVersion(pkg)), "\n")
}))
```
### Set up plotting conditions
```{r}
#set up plotting conditions
trt<-c("FASW","ABS","labrenzia","ruegeria","vibrio","ISOABC")
myPal<-c('grey60','black','#A6CEE3','#1F78B4','#B2DF8A','violet')
names(myPal) = trt

dpis<-c("-1","0","1","4","14","21")
shpPal<-c(0,15,11,13,5,1)
names(shpPal) = dpis

dpiPal<-c('black','gray30','seashell','seashell2','seashell3','seashell4')
names(dpiPal) <- c('-1','0','1','4','14','21')

theme_new<- theme_bw() + theme(panel.grid.major = element_line(color='gray90'), panel.grid.minor = element_line(color='gray90'), text= element_text(size = 15))
```

# Load and Manipulate Data

set paths
```{r}
path.in<-"../data/16S"
path.out<-"../output/16S"
```


```{r}
ps.exp<-readRDS(file.path(path.in,'ps.exp.samples.rds'))
ps.exp
```
fix taxa names
```{r}
ps.exp<-ps.exp %>%tax_fix()
```

## Sample sizes
```{r}
samps.sizes<-data.frame(sample_data(ps.exp)) %>% group_by(treatment, timepoint) %>% summarise(count=length(sample.names))
samps.sizes %>% arrange(treatment, timepoint,count)
```


## Normalize phyloseq object

```{r}
ps.exp.agg<-ps.exp %>% tax_agg("Species")
```

## clr norm
```{r}
ps.clr<-ps.exp.agg %>%  tax_transform(trans="clr")
```

## rel abund norm
```{r}
ps.relAbund<-ps.exp.agg %>% 
  tax_transform(trans="compositional") 
```

## rarefy data

```{r}
# 1. Extract OTU table and metadata
otu_mat <- as.data.frame(otu_table(ps.exp.agg))
metadata <- as(sample_data(ps.exp.agg), "data.frame")

# 2. Calculate rarefaction for each sample
rare_out <- rarecurve(otu_mat, step = 50, tidy = TRUE)

# 3. Merge rarefaction data with your sample metadata
rare_data <- rare_out %>%
    rename(SampleID = Site, Reads = Sample, Observed_ASVs = Species) %>%
    left_join(rownames_to_column(metadata, var = "SampleID"), by = "SampleID")

# 4. Plot using ggplot2, coloring by your group (e.g., TreatmentGroup)
ggplot(rare_data, aes(x = Reads, y = Observed_ASVs, group = SampleID, color = treatment)) +
  geom_line(linewidth = 0.8, alpha = 0.7) +
  theme_bw() +
  labs(
    x = "Number of Sequences",
    y = "Observed Richness (ASVs/OTUs)",
    color = "Sample Group"
  ) + 
  xlim(0,6000) + ylim(0,75)
```


```{r}
set.seed(200)
ps.exp.agg.reduced<-prune_samples(sample_sums(ps.exp.agg) >=4900, ps.exp.agg)
samps.sizes<-data.frame(sample_data(ps.exp.agg.reduced)) %>% group_by(treatment, timepoint) %>% summarise(count=length(sample.names))
samps.sizes %>% arrange(count)
```
```{r}
set.seed(1245)
ps.rare<- rarefy_even_depth(ps.exp.agg.reduced,rngseed = T)
ps.rare
```

# Priority Effect Analysis
Study design: 

Anemones were treated with ABS solutions and then incoculated with one of the following: 
- Isolate A - Labrenzia 
- Isolate B - Reugeria 
- Isolate C - Vibrio
- All three isolates 
- No isolates (ABS control)

There were also anemones that were held in FASW for the entire experiment

Priority effect logic:
- Single-strain treatments represent "founder effects"
- Does early establishment of one strain alter the final community composition compared to a null model (i.e., ABS or All three strains)?
- A strong priority effect = single-strain communities diverge from the
community treatment and resist convergence over time.

Evidence for priority effects: 
- Single strain inoculations remain compositionally distinct at day 21
- PREMANOVA: Significant Time X Treatment Effect
- 

## Beta-diversity analysis
Community composition over time 
- Remove FASW samples from analysis 
- Use rarefied data

### rarefied data
#### Ordination
```{r}
set.seed(1392)

ps.analysis<-ps.rare  %>% ps_filter(dpi != "-1") #%>%  ps_filter(treatment != "FASW") 

#reset factor level variables
sample_data(ps.analysis)$timepoint<-factor(sample_data(ps.analysis)$timepoint,levels=c("post_priming","post_ABS","1dpi","4dpi",'14dpi','21dpi'), ordered=T)
sample_data(ps.analysis)$dpi<-factor(sample_data(ps.analysis)$dpi,levels=c("-1","0","1","4","14","21"), ordered=T)
sample_data(ps.analysis)$treatment<-factor(sample_data(ps.analysis)$treatment,levels=c('FASW', 'ABS' ,'ISOABC', 'labrenzia', 'ruegeria', 'vibrio'), ordered=T)

#run ndms
nmds.mod <- ps.analysis %>% 
  dist_calc("bray") %>% 
  ord_calc(method = "NMDS") # 
#> Proportional min_prevalence given: 0.05 --> min 41/801 samples.
nmds.mod
nmds.mod@ord$stress
```

#####plot
```{r}
library(ggstar)
# n1<-nmds.mod %>%
#   ord_plot(shape=NA) + 
#   #scale_shape_manual(values=shpPal)+
#   theme_bw() +
#   stat_chull(
#     mapping = aes(colour = treatment, linetype = dpi,fill=treatment),
#     linewidth = 0.5, alpha = 0.25, show.legend = FALSE)+
#   geom_point(mapping=aes(color=treatment, shape=dpi),size=4,alpha=0.7)+
#   scale_fill_manual(values=myPal)+
#   scale_color_manual(values=myPal)+
#   scale_shape_manual(values=shpPal, guide=guide_legend(reverse=T))+
# n1

n1 <- nmds.mod %>%
ord_plot(shape=NA) +
theme_bw() +
stat_chull(
mapping = aes(colour = treatment, linetype = dpi, fill = treatment),
linewidth = 0.5, alpha = 0.25, show.legend = FALSE) +
geom_star(mapping = aes(color = treatment, fill = treatment, starshape = dpi),
size = 4, alpha = 0.7) +
  scale_fill_manual(values = myPal) +
scale_color_manual(values = myPal) +
scale_starshape_manual(
values = c(
"0" = 15,
"1"= 11,
"4"= 13,
"14" = 5,
"21" = 1
),
guide = guide_legend(reverse = TRUE)
)
n1
```
facet by timepoint

```{r}
dpi_labels <- c("1" = "1 dpi", "4" = "4 dpi", "14" = "14 dpi", "21" = "21 dpi", "0"="0 dpi","-1"="pre-inc")

n_s1 <- nmds.mod %>%
  ord_plot(shape=NA) + 
  scale_shape_manual(values=shpPal)+
  theme_bw() +
  stat_chull(
    mapping = aes(colour = treatment, linetype = dpi,fill=treatment),
    linewidth = 0.5, alpha = 0.25, show.legend = FALSE)+
  geom_point(mapping=aes(color=treatment, shape=dpi),size=4,alpha=0.7)+
  scale_fill_manual(values=myPal)+
  scale_color_manual(values=myPal)+
  scale_shape_manual(values=shpPal, guide=guide_legend(reverse=T))+
  facet_grid(~ dpi,
             labeller = labeller(dpi = dpi_labels)) +
 theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
n_s1
```
```{r}
n_s1 <- nmds.mod %>%
ord_plot(shape=NA) +
theme_bw() +
stat_chull(
mapping = aes(colour = treatment, linetype = dpi, fill = treatment),
linewidth = 0.5, alpha = 0.25, show.legend = FALSE) +
geom_star(mapping = aes(color = treatment, fill = treatment, starshape = dpi),
size = 4, alpha = 0.7) +
  scale_fill_manual(values = myPal) +
scale_color_manual(values = myPal) +
scale_starshape_manual(
values = c(
"0" = 15,
"1"= 11,
"4"= 13,
"14" = 5,
"21" = 1
),
guide = guide_legend(reverse = TRUE)
) + 
  facet_grid(~ dpi,
             labeller = labeller(dpi = dpi_labels)) + 
   theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
n_s1
```


#### PERMANOVA
```{r}
permanova<-ps.analysis %>% 
  dist_calc("bray") %>% 
  dist_permanova(
    variables = c("treatment","dpi","treatment:dpi"),
    n_perms = 9999, 
    seed = 1392, complete_cases = TRUE, verbose = "max"
  )
```

Significant interaction between treatment and time, suggests priority effects of microbial treatments
```{r}
print(permanova)
```
#### Pairwise PERMANOVA
```{r}
# Build distance matrix (samples as rows)
tmp.pw <- as.data.frame(otu_table(ps.analysis))
d.pw <- vegdist(tmp.pw, method = 'bray')

# Get metadata aligned to distance matrix sample order
mdf.pw <- data.frame(sample_data(ps.analysis))

# Run pairwise PERMANOVA across treatment groups
pw_results <- pairwise.adonis(d.pw, mdf.pw$treatment, p.adjust.m = "BH", perm = 999)
pw_results %>% arrange(p.adjusted)
```

#### Comparison to null model 
```{r}
library(dplyr)
library(tidyr)

metadata<-data.frame(sample_data(ps.analysis)) %>% 
          rownames_to_column("sample_id")

# Convert distance matrix to long format
bc_long<-as.matrix(dist_get(nmds.mod)) %>% 
  as.data.frame() %>% rownames_to_column("sample1") %>%
  pivot_longer(-sample1, names_to = "sample2", values_to = "bray_dist") %>% arrange(bray_dist,sample1)

# 2. Identify FASW DPI at 0 reference samples
fasw_ref_samples <- metadata %>%
  filter(treatment == "FASW", dpi == '0') %>%
  pull(sample_id)

# 3. For each sample, calculate its mean BC distance to all FASW DPI 0
mean_dist_to_ref <- bc_long %>%
  filter(sample2 %in% fasw_ref_samples,  # only distances to reference samples
         sample1 != sample2) %>%          # exclude self-comparisons
  group_by(sample1) %>%
  summarise(mean_dist_to_fasw_dpi0 = mean(bray_dist), .groups = "drop")

# 4. Join with metadata
result <- metadata %>%
  left_join(mean_dist_to_ref, by = c("sample_id"= "sample1")) %>% 
  mutate(dpi=forcats::fct_drop(dpi), treatment=fct_drop(treatment)) #%>% 
  #mutate(dpi=as.vector(dpi),treatment=as.vector(treatment))

str(result)
```


code here plots the average BC distance of each treatment group at each time point in comparison to T=0 FASW controls 

```{r}
bcd<-ggplot(result, aes(x=dpi, y=mean_dist_to_fasw_dpi0,fill=treatment, color=treatment)) + 
  geom_boxplot(alpha=0.5, outlier.shape=NA,position = position_dodge2(preserve="single")) + 
  geom_jitter(position = position_dodge2(0.75,preserve="single"),size=1.5)+
  ylab('Avg. Bray Curtis distance to FASW (t=0 dpi)')+
  scale_fill_manual(values = myPal) +
  scale_color_manual(values = myPal)+
  theme_new +
  theme(title = element_text(size = 12))+
  theme(legend.title = element_blank(),
        legend.text = element_text(size=9))
bcd
```

##### GLM - beta regression 

```{r}
result
```
```{r}
#fix ordered vectors
result <- metadata %>%
  left_join(mean_dist_to_ref, by = c("sample_id"= "sample1")) %>% 
  mutate(dpi=forcats::fct_drop(dpi), treatment=fct_drop(treatment)) %>% 
  mutate(dpi=as.vector(dpi),treatment=as.vector(treatment))
```

How does distance from a reference change over time?
```{r}
library(glmmTMB)

#run a glm using beta regression but acknowlding non independence limitation 
model_tmb <- glmmTMB(mean_dist_to_fasw_dpi0 ~ treatment * dpi + (1|sample.names),
                     data   = result,
                     family = beta_family(link = "logit")) 
summary(model_tmb)
```

```{r}
car::Anova(model_tmb, type = "II")
```
```{r}
# test interaction term
model_tmb_noint <- glmmTMB(mean_dist_to_fasw_dpi0 ~ treatment + dpi + (1 | sample.names),
                            data   = result,
                            family = beta_family(link = "logit"))

anova(model_tmb_noint, model_tmb)  # LRT for interaction
```

```{r}
emmeans(model_tmb,  ~ treatment * dpi  , adjust = "BH")
```
###### Check model assumptions

```{r}
library(DHARMa)

# Simulate residuals 
par(mfrow=c(2,2))
sim_res <- simulateResiduals(model_tmb, plot = TRUE)
testDispersion(sim_res) #over disperstion vs under dispersion 
testZeroInflation(sim_res) # excess 0s?
testOutliers(sim_res) #outliers? 
```

##### Betadispersion
"do some treatment × dpi combinations have more variable communities than others?"
```{r}
dist<-dist_get(nmds.mod)
groups<-interaction(nmds.mod@sam_data$treatment, nmds.mod@sam_data$dpi)
bd <- betadisper(dist, groups)
anova(bd)          # overall test
permutest(bd)      # permutation-based, more robust
```

```{r}
#run pairwise comparison 
tukey_result <- TukeyHSD(bd)
# Extract the pairwise comparisons table
tukey_table <- as.data.frame(tukey_result$group)
# Sort by p-value lowest first
tukey_table <- tukey_table[order(tukey_table$`p adj`), ]
tukey_table
```
```{r}
plot(bd)
boxplot(bd, las = 2)
```


# FAVA analysis 
From https://maikemorrison.github.io/FAVA/
Goal of FAVA is to quantify compositional variability in a dataset across treatments using an FST appraoch
https://maikemorrison.github.io/FAVA/articles/microbiome_tutorial.html

## prep data
```{r}
library(FAVA)
#select data
ps.exp.ra<-microbiome::transform(ps.exp, "compositional")
ps.analysis<-ps.exp.ra %>% ps_filter(dpi != "-1")

my_data<-relab_phyloseq(ps.analysis) %>% 
  dplyr::select(-is.neg,-Index,-LibrarySize,-sample_or_control, -sample.names3,-sample.names2,-DNA_conc_ng_ul) %>%
  mutate(treatment=as.vector(treatment), dpi=as.numeric(as.vector(dpi))) %>% arrange(dpi)

str(my_data)
```
```{r}
# Confirm that your samples each sum to 1
# if columns 4 through 10 contain the relevant categories
# and columns 1, 2, and 3 contain metadata
rowSums(my_data[,c(5:ncol(my_data))])
```
## check matrixs match
```{r}
# specify tree
tree <- phy_tree(ps.analysis)

#get phylogenetic distances
distance_matrix <- ape::cophenetic.phylo(tree)

#similarity matrix
similarity_matrix = exp(-distance_matrix)

# alternative similarity matrices:
# similarity_matrix = 1/(distance_matrix + 1)
# similarity_matrix = 1 - distance_matrix/max(distance_matrix)

# Get the names of the species in your relative abundance matrix
species_order <- colnames(my_data[,c(5:ncol(my_data))])

# Confirm that the entries of the similarity matrix
# correspond to relative abundance matrix
all(species_order == colnames(similarity_matrix))
all(species_order == rownames(similarity_matrix))

# If they do not, you can re-order the rows and columns of
# your similarity matrix to match your data:
# similarity_matrix_reordered = similarity_matrix[species_order, species_order]

# confirm that all diagonal elements are still 1
diag(similarity_matrix)
```
## run analysis
### visualize relative abudnances 
```{r}
set.seed(1)
nr<-length(5:ncol(my_data))
species_palette = viridis::turbo(nr)[sample(1:nr)] %>%
  `names<-`(colnames(my_data)[-c(1:5)])

# Make a ggplot2 stacked bar plot
plot_relabund(my_data,
              group = "treatment",
              arrange = "vertical",
              K = nr) +
# Specify a custom color scheme
  ggplot2::scale_color_manual(values = species_palette) +
  ggplot2::scale_fill_manual(values = species_palette)
```
### compute unweighted FAVA
```{r}
fava.out<-fava(relab_matrix = my_data,
     group = c("treatment","dpi"),
     K = nr, normalized = T)
```
#### plot FAVA
higher values of FAVA mean treatment is more variable at that time point vs. lower values of FAVA. 

```{r}
fa<-ggplot(fava.out, aes(x = dpi, y = FAVA,color=treatment)) + 
  geom_point(size = 3) + 
    geom_line()+
  scale_color_manual(values=myPal)+
  theme_new
fa
```
FAVA tells us that the ABS samples at the start of the experiment are most variable across individuals, while toward the end of the experiment vibrio has the most variation in the FAVA score. 
- ABS consistent with past work; ABS induces disorganization of community during initial time points with highest variation on 4 dpi. Then the composition variation of the ABS samples converges as the community becomes more similar. 
- FASW stays similar across experiment 
- for most bac inoculation treatments FAVA score increases 4 DPI but then levels off, except vibrio
- Vibrio FAVA score increases and keeps increasing suggesting Vibrio had the longest impact on the variation of the community in the experiment

#### bootstrap
change this to include similarity matrix as well 

To determine if these differences in variability levels are statistically significant, we employ bootstrapping, a computational method used to generate a null distribution against which to test statistical hypotheses.
```{r}
#bootstrap_out <- bootstrap_fava(relab_matrix = my_data,
#                               n_replicates = 500,  
#                               seed = 3,
#                               group = c("treatment","dpi"),
#                               K = nr)
```
get pvalues
```{r}
#str(bootstrap_out, max.level = 1)
#bootstrap_out$P_values %>% arrange(P_value)
```

plot results
```{r}
#bootstrap_out$bootstrap_distribution_plot
```




# ANCOM-BC2 Analysis

Runs a differential abundance analysis comparing the abundance of each ASV between the ABS samples and the treatment samples across each timepoint (1, 4, 14 and 21 dpi)

## Set up analysis
```{r}
library(ANCOMBC)
ps.analysis<-ps.exp %>% ps_filter(dpi != "-1")
treatments_ancom <- c("labrenzia", "ruegeria", "vibrio", "ISOABC")
timepoints_ancom <- c("1dpi", "4dpi", "14dpi", "21dpi")
```

## run analysis 
compares each treatment to ABS conditions at each time point 
```{r}
ancom_results <- list()

for (tp in timepoints_ancom) { #loops timepoints
  for (trt in treatments_ancom) { #loops treatments
    
     # subset to ABS + focal treatment
    ps.sub <- ps.analysis %>%
      ps_filter(treatment %in% c("ABS", trt)) %>%
      ps_filter(timepoint == tp) %>%
      tax_filter(min_prevalence = 0.1)  # remove taxa present in < 10% of samples
    
    # skip small groups
    grp_sizes <- table(sample_data(ps.sub)$treatment)
    if (any(grp_sizes < 2)) next
    
    # run ANCOM-BC2
    res <- ancombc2(
      data          = ps.sub,
      tax_level     = NULL,
      fix_formula   = "treatment",
      p_adj_method  = "BH",
      group         = "treatment",
      struc_zero    = F, 
      neg_lb        = TRUE,
      verbose       = FALSE,
      pseudo        = TRUE,
      pseudo_sens   = TRUE
    )
    
    # rename treatment-specific columns to generic names
    res_df <- res$res %>%
      dplyr::select(
        taxon,
        lfc        = paste0("lfc_treatment", trt),
        se         = paste0("se_treatment", trt),
        W          = paste0("W_treatment", trt),
        p          = paste0("p_treatment", trt),
        q          = paste0("q_treatment", trt),
        diff       = paste0("diff_treatment", trt),
        passed_ss  = paste0("passed_ss_treatment", trt)
      ) %>%
      mutate(timepoint  = tp,
             comparison = paste0(trt, "_vs_ABS"))
    
    ancom_results[[paste(tp, trt)]] <- res_df
  }
}

ancom_results_df <- bind_rows(ancom_results)

# significant hits
ancom_sig <- ancom_results_df %>% filter(diff == TRUE)
ancom_sig %>% arrange(comparison, timepoint, q)
```

## Visualize 
```{r}
# join taxonomy to get genus names instead of ASV IDs
tax_df <- tax_table(ps.analysis) %>% 
  as.data.frame() %>% 
  rownames_to_column("taxon") %>%
  dplyr::select(taxon, Species)

ancom_sig_2 <- ancom_sig %>%
  left_join(tax_df, by = "taxon") %>%
  mutate(
    sig_label = case_when(
      q < 0.001 ~ "***",
      q < 0.01  ~ "**",
      q < 0.05  ~ "*",
      TRUE      ~ ""
    ),
    direction = ifelse(lfc > 0, "Enriched vs ABS", "Depleted vs ABS")
  ) %>% 
   mutate(Species_short = gsub("\\(.*\\)", "", Species),   # drop accession
         Species_short = gsub("_A$|_B$", "", Species_short),  # drop _A/_B suffixes if unwanted
         Species_short = trimws(Species_short))
  
```

```{r}
#reorder species, treatments and timepoints
sig_species <- unique(ancom_sig_2$Species_short)
priority_species<-c("Roseibium aggregatum","Epibacterium scottomollicae","Vibrio alginolyticus")

other_species  <- sort(setdiff(unique(ancom_sig_2$Species_short), priority_species))
ancom_sig_2   <- ancom_sig_2 %>%
  mutate(Species_short = factor(Species_short, levels = rev(c(priority_species, other_species)))) %>% 
  mutate(timepoint=factor(timepoint, levels=c('1dpi','4dpi','14dpi','21dpi'),ordered=T),
         comparison=factor(comparison,levels=c("ISOABC_vs_ABS","labrenzia_vs_ABS", "ruegeria_vs_ABS",  "vibrio_vs_ABS"),ordered=T)) %>% 
  mutate(comparison2=recode(comparison, 'ISOABC_vs_ABS'='ISOABC','labrenzia_vs_ABS'='Labrenzia','ruegeria_vs_ABS'='Ruegeria','vibrio_vs_ABS'='Vibrio'))
```


```{r}
diff_taxa<-ggplot(ancom_sig_2, aes(x = timepoint, y = Species_short, fill = lfc)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = sig_label), size = 4, vjust = 0.75) +
  facet_grid(~ comparison2) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, name = "Log Fold\nChange vs ABS"
  ) +
  scale_x_discrete(limits = c("1dpi", "4dpi", "14dpi", "21dpi")) +
  theme_bw() +
  theme(
    axis.text.y      = element_text(size = 8, face = "italic"),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    strip.text       = element_text(face = "bold"),
    strip.background = element_blank(),
    panel.grid       = element_blank()
  ) +
  labs(x = "Timepoint", y = "Species",
       title = "Differentially Abundant Taxa vs ABS")
diff_taxa
```
- This is plotting the deferentially abundant taxa as logfold 2 change between treatment and ABS where red is higher in treatment and blue is higher in ABS samples for each timepoint
- ISOABC vs. ABS: both labrenzia (Roseibium) and ruegeria (Epibacterium) came up as being diff abund at start of experiment, do not see diff abundance of vibrio --> labrenzia and ruegeria did a good job at outcompeting vibrio for space 
- Labrenzia -> diff abundant during 1, 4 and 14 dpi
- Ruegeria -> diff abundnace at 1 dpi, nr of taxa that diff abudnnace in tratment at end of experiment, including: 
  - Alcanivoraz 
  - Bdellovibrionaceae family
  - Hoeflea 
  - Mamelilla alba 
  - Marcaulis maris 
- Vibrio -> diff abundance at 1 and 4 dpi
  - at end of expeirment, altermonas portus diff abundance 
- In all inoculation vs. abs samples, acorss most timepoints see sig depletion of Shimia species, including shimia aestuarii, s. haliotis

  

```{r}
#get sig species names
sig_species_full <- unique(ancom_sig_2$Species)
priority_species<-c("Roseibium aggregatum_A(RS_GCF_001999245_1)","Epibacterium scottomollicae(RS_GCF_003003215_1)","Vibrio alginolyticus(RS_GCF_000354175_2)")

#extract rel abundance for sig. species
relabund_heat<- psmelt(ps_get(ps.relAbund)) %>%
  filter(Species %in% sig_species_full) %>%
  filter(timepoint %in% c("1dpi", "4dpi", "14dpi", "21dpi")) %>%
  filter(treatment %in% c('ABS' ,'ISOABC', 'labrenzia' ,'ruegeria' ,'vibrio')) %>% 
  group_by(Species, treatment, timepoint) %>%
  summarise(mean_rel_abund = mean(Abundance, na.rm = TRUE), .groups = "drop")

# join significance labels from ancom_sig
sig_overlay <- ancom_sig_2 %>%
  mutate(treatment = str_remove(comparison, "_vs_ABS")) %>% 
  dplyr::select(Species, timepoint, treatment, sig_label)

#join sig levels and rel abundance 
relabund_heat.sig_overlay <- relabund_heat %>%
  left_join(sig_overlay, by = c("Species", "timepoint", "treatment")) %>%
  mutate(sig_label = replace_na(sig_label, ""))

# z-score each genus across all treatment/timepoint combinations
df.heat <- relabund_heat.sig_overlay %>%
  group_by(Species) %>%
  mutate(zscore = scale(mean_rel_abund)[, 1]) %>%
  ungroup()

# fix ordering of species, treatments, and get short species names 
df.heat<-df.heat %>%  #get short names
         mutate(Species_short = gsub("\\(.*\\)", "", Species),   # drop accession
         Species_short = gsub("_A$|_B$", "", Species_short),  # drop _A/_B suffixes if unwanted
         Species_short = trimws(Species_short))

# reorder species: priority species first, then remaining alphabetically
priority.sp<-c("Epibacterium scottomollicae","Roseibium aggregatum","Vibrio alginolyticus")
other.sp<-sort(setdiff(unique(df.heat$Species_short), priority.sp))

# reorder things
df.heat <-df.heat %>% 
          mutate(Species_short=factor(Species_short,levels=rev(c(priority.sp,other.sp)))) %>% 
          mutate(timepoint=factor(timepoint, levels=c('1dpi','4dpi','14dpi','21dpi')),
                 treatment=factor(treatment,levels=c('ABS' ,'ISOABC', 'labrenzia' ,'ruegeria' ,'vibrio')))

ggplot(df.heat, aes(x = timepoint, y = Species_short, fill = mean_rel_abund)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = sig_label), size = 3.5, vjust = 0.75, color = "black") +
  facet_wrap(~ treatment, nrow = 1) +
  scale_fill_gradientn(
    colours  = c("white", "#FED976", "#FD8D3C", "#E31A1C", "#800026"),
    name     = "Mean\nRelative\nAbundance",
    labels   = scales::percent_format(accuracy = 0.1),
    na.value = "grey90"
  ) +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    axis.text.y      = element_text(size = 8, face = "italic"),
    strip.text       = element_text(face = "bold"),
    strip.background = element_blank(),
    panel.grid       = element_blank(),
    legend.title     = element_text(size = 9)
  ) +
  labs(x = "Timepoint", y = "Species",
       title    = "Relative Abundance of Differentially Abundant Taxa",
       subtitle = "Asterisks indicate significance Treatment vs ABS (ANCOM-BC2)")
```
In the entire experiment, there were 30 bacterial species / ASVs that were sig different between treatment and time point comparisons. 



# Barplot
```{r}
# set taxon order: inoculants first, then remaining by abundance
all_taxa      <- tax_table(ps.relAbund)[, "Species"] %>%
  as.data.frame() %>% pull(Species) %>% unique()
other_taxa    <- setdiff(all_taxa, priority_species)
tax_order     <- c(priority_species, other_taxa)

# clean facet labels
trt_labels <- c(
  "ABS"       = "ABS",
  "labrenzia" = "Labrenzia",
  "ruegeria"  = "Ruegeria",
  "vibrio"    = "Vibrio",
  "ISOABC"    = "All Three"
)
tp_labels <- c(
  "post_priming" = "Pre-ABS",
  "post_ABS"     = "Post-ABS",
  "1dpi"         = "1 dpi",
  "4dpi"         = "4 dpi",
  "14dpi"        = "14 dpi",
  "21dpi"        = "21 dpi"
)
```

```{r}
library(ggh4x)

bar<-ps.relAbund %>%
  ps_filter(timepoint %in% c("1dpi", "4dpi", "14dpi", "21dpi")) %>%
  ps_mutate(
    dpi      = as.factor(dpi),
    timepoint = factor(timepoint, levels = c("1dpi", "4dpi", "14dpi", "21dpi"))
  ) %>%
  tax_sort(by = sum, at = "Species") %>%
  comp_barplot(tax_level = 'Species', n_taxa = 30,
               merge_other = FALSE,
               other_name  = "Other",
               x           = 'sample.names',
               tax_order   = tax_order) +
  facet_nested(.~treatment + timepoint,
               scales   = 'free', space = "free",
               labeller = labeller(treatment = trt_labels,
                                   timepoint = tp_labels)) +
  theme(
    #axis.text.y      = element_blank(),
    #axis.ticks.y     = element_blank(),
    strip.text       = element_text(size = 9, face = "bold"),
    #strip.background = element_blank(),
    panel.spacing    = unit(0.2, "lines")
  ) +
  ylab('Relative Abundance') + xlab('')

bar
```


# Figures
```{r}
library(patchwork)
```

## Main Figure

```{r}
bcd.2<-bcd + theme(legend.position="none") + theme(axis.title = element_text(size=12)) + theme(legend.title  = element_blank())

n1.2<-n1+ theme(axis.title = element_text(size=12))+ theme(legend.title = element_blank()) 

(n1.2 + bcd.2) / diff_taxa +  plot_layout(nrow=2, widths=c(1,1)) + plot_annotation(tag_levels='A')

ggsave(file.path(path.out,'FigX_16S.svg'),width=14, height=8,units = "in" )
```
```{r}
bcd.2
```


```{r}
bcd.2<-bcd + theme(legend.position="none") + theme(axis.title = element_text(size=12)) + theme(legend.title  = element_blank())
bar.2<- bar + theme(legend.position="bottom") +  theme(axis.title = element_text(size=12))+ theme(legend.title = element_blank())
n1.2<-n1+ theme(axis.title = element_text(size=12))+ theme(legend.title = element_blank()) 
bar.2 + n1.2 + bcd.2 + plot_layout(nrow=2, widths=c(1.5,1,0.5)) + plot_annotation(tag_levels='A')

ggsave(file.path(path.out,'FigX_16s.svg'),width=14, height=8,units = "in" )
```
```{r}
n1.2
ggsave(file.path(path.out,'ordination.pdf'),units = 'in',width = 8,height=7)
```

```{r}
bar
ggsave(file.path(path.out,'barplot.pdf'),units = 'in',width = 16,height=7)
```


## Supplemental figures

### nmds faceted plot
```{r}
n_s1
ggsave(file.path(path.out,'FigSX_nmds_facets.pdf'),units = 'in',width = 8,height=4)
```

### FAVA analysis 
```{r}
fa
ggsave(file.path(path.out,'FigSX_fava.pdf'),units = 'in',width = 5,height=3.5)
```

### Diff Abund taxa
```{r}
diff_taxa + ggtitle('ANCOM Sig Taxa: Trt vs. ABS')
ggsave(file.path(path.out,'FigSX_diffABund.pdf'),units = 'in',width = 10,height=5)
```
