---
title: "RNASeq data processing"
author: "Hailey Hatch & Maggie Sogin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
---

# Introduction

This notebook processes metatranscriptomic (RNA-seq) data from an *Aiptasia* bacterial inoculation experiment. *Aiptasia* were first treated with antibiotics to reduce their native microbiome, then inoculated with one of three bacterial symbionts to examine changes in host, algal symbiont, and bacterial gene expression.

The script below aims to process data from RNA-Seq experiment and includes the following: 

1. Load in data 
2. Get Counts from featurecounts (saved) 
3. Get TPM Counts (not saved)
4. QC plots
5. Clean and split tables, save by organism
6. Prepare annotations

At the end of this code, we save the following files for downstream analysis:
- cleaned feature counts table per species
- processed annotation files for aiptasia, bacteria and symbiodinium

**Previous pipeline:** Raw reads were processed with fastp (quality trimming) → SortMeRNA (rRNA removal) → HISAT2 (alignment to a combined reference genome containing *Aiptasia* host, *Symbiodiniaceae* symbiont, *Labrenzia*, *Ruegeria*, and *Vibrio*) → featureCounts (gene-level quantification).


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,           # Show code in the output
  warning = FALSE,       # Hide warning messages
  message = FALSE,       # Hide package loading messages
  fig.width = 10,        # Default figure width (inches)
  fig.height = 6,        # Default figure height (inches)
  fig.align = "center"   # Center all figures
)
```


# 1. Load Libraries & Define Constants

```{r load-libraries}
library(readxl)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(matrixStats)
library(ggrepel)
library(patchwork)
library(UpSetR)
library(stringr)
library(topGO)
library(vegan)
library(dplyr)
library("RColorBrewer")
library(scales)
```

```{r define-constants}
palette <- c(
  "FASW" = 'grey60',
  "ABS"     = 'black',
  "Labrenzia"    = '#A6CEE3',
  "Ruegeria"    = '#1F78B4',
  "Vibrio"    = '#B2DF8A',
  "multi"  = 'violet')
treatment_levels <- c(
  "control" = "FASW", 
  "ABS" = "ABS", 
  "ISOA" = "Labrenzia", 
  "ISOB" = "Ruegeria", 
  "ISOC" = "Vibrio", 
  "ISOABC" = "multi")

path.out<-"../../output/transcriptomics"
path.in<-"../../data/transcriptomics"
```


# 2. Load All Data Files

```{r load-all-data}
# featureCounts gene count matrix (6 annotation cols + 48 samples)
counts_raw <- read.delim(file.path(path.in,"counts.tsv"), comment.char = "#", check.names = FALSE)

# featureCounts assignment summary
counts_summary <- read.delim(file.path(path.in,"counts.tsv.summary"), check.names = FALSE)

# HISAT2 alignment rates
hisat2_sum <- read.delim(file.path(path.in,"hisat2_alignment_summary.tsv"))

# Per-reference organism read counts (aip, sym, lab, rue, vib)
organism_counts <- read.delim(file.path(path.in,"per_reference_counts.tsv"))

# SortMeRNA rRNA removal summary
sortme_sum <- read.delim(file.path(path.in,"sortmerna_summary.tsv"))

# Fastp trimming summary
fastp_sum <- read.delim(file.path(path.in,"fastp_summary_per_sample.tsv"), check.names = FALSE)

# Aiptasia gene annotations (SwissProt/TrEMBL BLAST, GO, KEGG)
aip_annotate <- read.delim(file.path(path.in,"Aiptasia_annotation_file.csv"))
length(aip_annotate$Protein_ID)
```


# 3. Build Sample Metadata

Create the `colData` data frame that DESeq2 requires. We'll parse treatment and timepoint directly from the sample names so it's fully reproducible.

```{r build-metadata}
# Step 1: Make function to clean sample names
clean_sample_name <- function(x) {gsub("_S\\d+.*$", "", x)}

# Step 2: Get sample column names from the counts data
colnames(counts_raw) <- clean_sample_name(basename(colnames(counts_raw)))
annotation_cols <- c("Geneid", "Chr", "Start", "End", "Strand", "Length")
sample_names <- colnames(counts_raw)[!colnames(counts_raw) %in% annotation_cols]

# Step 3: Parse treatment and timepoint from the clean names
treatment <- str_extract(sample_names, "ISOABC|ISOA|ISOB|ISOC|ABS|control")
timepoint <- str_extract(sample_names, "\\d+dpi")

# Step 4: Create the colData data frame
colData <- data.frame(row.names = sample_names, treatment = treatment, timepoint = timepoint)

# Step 5: Convert treatment and timepoint to factors with your defined levels
colData$treatment <- factor(treatment_levels[as.character(colData$treatment)], 
                            levels = treatment_levels)
colData$timepoint <- factor(colData$timepoint, levels = c("1dpi", "21dpi"))

stopifnot(all(table(colData$treatment, colData$timepoint) == 4))

# Make metadata by turning rownames into a regular column
metadata <- colData %>% rownames_to_column("sample")

# save colData

write.csv(colData, file.path(path.out,'colData.csv'))
```


# 4. QC Summary Plots

Visualize pipeline quality metrics to identify any problematic samples before DESeq2.

## 4.1 Fastp — Read Quality & Trimming

```{r plot-fastp}
plot_QC_fastp <- fastp_sum %>%
  mutate(sample = clean_sample_name(sample)) %>%
  left_join(metadata, by = "sample") %>%
  filter(!is.na(treatment)) %>%
  arrange(treatment, timepoint) %>%
  mutate(sample = fct_inorder(sample)) %>%
  dplyr::select(sample, treatment, timepoint, reads_before, reads_after) %>%
  pivot_longer(cols = c(reads_before, reads_after),
               names_to = "stage", values_to = "reads") %>%
  mutate(stage = factor(stage, levels = c("reads_before", "reads_after"),
                        labels = c("Raw", "After Trimming"))) %>%
  ggplot(aes(x = sample, y = reads / 1e6, fill = stage)) +
  geom_col(position = "dodge") +
  facet_wrap(~ timepoint, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Fastp: Read Counts Before & After Trimming",
       y = "Reads (millions)", x = NULL)
plot_QC_fastp
# No extreme trimming; all samples ~125-225M reads
```
## 4.2 SortMeRNA — rRNA Removal Rates

```{r plot-sortmerna}
plot_QC_sortmerna <- sortme_sum %>%
  mutate(sample = clean_sample_name(sample)) %>%
  left_join(metadata, by = "sample") %>%
  arrange(treatment, timepoint) %>%
  mutate(sample = fct_inorder(sample)) %>%
  ggplot(aes(x = sample, y = rRNA_pct, fill = treatment)) +
  geom_col() +
  facet_wrap(~ timepoint, scales = "free_x") +
  scale_fill_manual(values = palette) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "SortMeRNA: rRNA Removal Rates", y = "rRNA (%)", x = NULL)
plot_QC_sortmerna
# ~5-20% rRNA; 21dpi slightly higher on average
```
## 4.3 HISAT2 — Alignment Rates

```{r plot-hisat2}
plot_QC_hisat2 <- hisat2_sum %>%
  mutate(sample = clean_sample_name(sample)) %>%
  left_join(metadata, by = "sample") %>%
  arrange(treatment, timepoint) %>%
  mutate(sample = fct_inorder(sample)) %>%
  ggplot(aes(x = sample, y = `overall_alignment_rate...`, fill = treatment)) +
  geom_col() +
  facet_wrap(~ timepoint, scales = "free_x") +
  scale_fill_manual(values = palette) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "HISAT2: Alignment Rates", y = "Alignment Rate (%)", x = NULL)
plot_QC_hisat2
# Alignment rates 50-75%
```
## 4.4  % Library Mapped to Each Genome

```{r fig-s1-pct-mapped, fig.width=12, fig.height=6}
# Compute % of total mapped reads assigned to each organism per sample
figS1_data <- organism_counts %>%
  mutate(sample = clean_sample_name(sample)) %>%
  left_join(metadata, by = "sample") %>%
  mutate(total = aip + sym + lab + rue + vib) %>%
  pivot_longer(cols = c(aip, sym, lab, rue, vib),
               names_to = "organism", values_to = "reads") %>%
  mutate(pct_mapped = 100 * reads / total,
         organism = factor(organism,
                           levels = c("aip", "sym", "lab", "rue", "vib"),
                           labels = c("Host", "Symbiodinium", "Labrenzia", "Ruegeria", "Vibrio")),
         treatment = factor(treatment, levels = treatment_levels))

plot_figS1 <- ggplot(figS1_data, aes(x = treatment, y = pct_mapped, color = organism)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 2.5, alpha = 0.8) +
  facet_grid(organism ~ timepoint, scales = "free_y") +
  scale_color_manual(values = c("Host" = "#1b9e77", "Symbiodinium" = "#d95f02",
                                "Labrenzia" = "#A6CEE3", "Ruegeria" = "#1F78B4",
                                "Vibrio" = "#B2DF8A")) +
  ylim(0, NA) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        strip.text = element_text(face = "bold")) +
  labs(title = "Fig S1 — % Library Mapped to Each Genome",
       y = "% of reads mapped to genome", x = NULL)
plot_figS1


ggsave(file.path(path.out,"Fig_pct_mapped_by_genome.pdf"), plot_figS1, width = 12, height = 6)
ggsave(file.path(path.out,"Fig_pct_mapped_by_genome.png"), plot_figS1, width = 12, height = 6, dpi = 300)
```
## 4.5 featureCounts — Read Assignment Summary

```{r plot-featurecounts}
colnames(counts_summary) <- clean_sample_name(basename(colnames(counts_summary)))

fc_data <- counts_summary %>%
  pivot_longer(cols = -Status, names_to = "sample", values_to = "reads") %>%
  filter(reads > 0) %>%
  left_join(metadata, by = "sample") %>%
  arrange(treatment, timepoint) %>%
  mutate(sample = fct_inorder(sample))

plot_QC_featurecounts <- ggplot(fc_data, aes(x = sample, y = reads, fill = Status)) +
  geom_col(position = "fill") +
  facet_wrap(~ timepoint, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "featureCounts: Read Assignment", y = "Proportion", x = NULL)
plot_QC_featurecounts
# Assigned reads ~25% — low but expected for metatranscriptomic reference
```

## 4.6 featureCounts — Library Size

```{r plot-library-size}
plot_QC_library_size <- fc_data %>%
  filter(Status == "Assigned") %>%
  ggplot(aes(x = sample, y = reads / 1e6, fill = treatment)) +
  geom_col() +
  facet_wrap(~ timepoint, scales = "free_x") +
  scale_fill_manual(values = palette) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Library Size (Assigned Reads)", y = "Reads (millions)", x = NULL)
plot_QC_library_size
# ~20-40M reads assigned to combined reference per sample
```
## 4.7 Save QC Plots

```{r save-qc-plots}
qc_plots <- list(
  "plot_QC_fastp"         = plot_QC_fastp,
  "plot_QC_sortmerna"     = plot_QC_sortmerna,
  "plot_QC_hisat2"        = plot_QC_hisat2,
  "plot_QC_featurecounts" = plot_QC_featurecounts,
  "plot_QC_library_size"  = plot_QC_library_size)

for (name in names(qc_plots)) {
  ggsave(filename = file.path(path.out,paste0(name, ".pdf")),
         plot = qc_plots[[name]],
         width = 10, height = 6)
  
  ggsave(filename = file.path(path.out,paste0(name, ".png")),
         plot = qc_plots[[name]],
         width = 10, height = 6, dpi = 300)
}
```

---

# 5. Clean and Prepare the Count Matrix

Transform the raw featureCounts output into a clean integer count matrix that DESeq2 can use.

```{r clean-counts}
# Save gene lengths for TPM calculation later
gene_lengths <- counts_raw %>% 
                dplyr::select(Geneid, Length) %>% column_to_rownames("Geneid")

counts_matrix <- counts_raw %>%
  dplyr::select(-Chr, -Start, -End, -Strand, -Length) %>%
  column_to_rownames("Geneid") %>%
  as.matrix()
counts_matrix <- round(counts_matrix)

all(colnames(counts_matrix) == rownames(colData))

dim(counts_matrix)
head(counts_matrix)
```

## 5.1 Compute TPM Matrix

TPM (Transcripts Per Million) normalizes for both gene length and sequencing depth,
making expression values comparable across samples AND genes

TPM formula: for each gene in each sample,
1. Divide raw count by gene length in kb → RPK (reads per kilobase)
2. Sum all RPKs in that sample → scaling factor
3. Divide each RPK by (scaling factor / 1e6) → TPM

```{r compute-tpm}
all(rownames(gene_lengths) == rownames(counts_matrix))

gene_lengths_kb <- gene_lengths$Length / 1000

rpk_matrix <- sweep(counts_matrix, 1, gene_lengths_kb, "/")

scaling_factors <- colSums(rpk_matrix)

tpm_matrix <- sweep(rpk_matrix, 2, scaling_factors / 1e6, "/")

tpm_aip <- tpm_matrix[startsWith(rownames(tpm_matrix), "aip_"), ]
tpm_sym <- tpm_matrix[startsWith(rownames(tpm_matrix), "sym_"), ]
tpm_lab <- tpm_matrix[startsWith(rownames(tpm_matrix), "lab_"), ]
tpm_rue <- tpm_matrix[startsWith(rownames(tpm_matrix), "rue_"), ]
tpm_vib <- tpm_matrix[startsWith(rownames(tpm_matrix), "vib_"), ]

dim(tpm_aip)
```

---

# 6. Split Counts by Organism

Analyzing organisms separately because:
- Library size normalization should be computed within an organism
- The biological questions differ per organism
- Read depths vary (host: millions, bacteria: thousands)

```{r split-by-organism}
counts_aip <- counts_matrix[startsWith(rownames(counts_matrix), "aip_"), ]
counts_sym <- counts_matrix[startsWith(rownames(counts_matrix), "sym_"), ]
counts_lab <- counts_matrix[startsWith(rownames(counts_matrix), "lab_"), ]
counts_rue <- counts_matrix[startsWith(rownames(counts_matrix), "rue_"), ]
counts_vib <- counts_matrix[startsWith(rownames(counts_matrix), "vib_"), ]

# Verify splits include all counts
nrow(counts_aip) + nrow(counts_sym) + nrow(counts_lab) + nrow(counts_rue) + nrow(counts_vib) == nrow(counts_matrix)
```

Gene filtering
- Below calculate the Nr of genes passing the filter, a gene is included in the filter if it has at least 10 reads in 4 samples across the entire dataset (Ref: deseq manual )

```{r organism-summary-stats}
organism_list <- list(
  aip = counts_aip, sym = counts_sym,
  lab = counts_lab, rue = counts_rue, vib = counts_vib)

organism_summary <- data.frame(
  organism = names(organism_list),
  total_genes = sapply(organism_list, nrow),
  median_counts_per_sample = sapply(organism_list, function(x) median(colSums(x))), #median nr. reads mapping to each sample
  genes_all_zero = sapply(organism_list, function(x) sum(rowSums(x) == 0)), #nr genes no expression data at all 
  genes_passing_filter = sapply(organism_list, function(x) sum(rowSums(x >= 10) >= 4)) # at least 10 counts in 4 samps
)
organism_summary
# Few bacterial genes passed the filter
```

```{r save-organism-counts}
for (name in names(organism_list)) {
  write.csv(organism_list[[name]],
            file = file.path(path.out, paste0("counts_", name, ".csv")),
            quote = FALSE, row.names = TRUE)
}
```

# 7. Prepare Annotation Data

The annotation file has multiple transcript isoforms per gene (e.g., g1.t1 and g1.t2 are isoforms of gene g1). We need to:

1. Extract the gene-level ID
2. Match it to the count matrix format (with `aip_` prefix)
3. Deduplicate to one annotation per gene 

## 7.1 Parse Aiptasia annotations
```{r prepare-annotations}
# Step 1: Extract gene ID from Protein_ID
aip_annotate <- aip_annotate %>%
  mutate(gene_id = paste0("aip_", sub("\\..*$", "", Protein_ID)))
aip_annotate_full<-aip_annotate

# Step 2: Extract a short protein name from the SwissProt BLAST hit column
aip_annotate <- aip_annotate %>%
  mutate(protein_name = str_extract(`BLAST_hit..SwissProt.`, "(?<=\\|)[^|]+(?=_)"))

# Step 3: Deduplicate — keep one row per gene_id
# aip_annotate <- aip_annotate %>%
#   group_by(gene_id) %>%
#   slice_min(`E.value..SwissProt.`, n = 1, with_ties = FALSE, na.rm=T) %>%
#   ungroup()

aip_annotate <- aip_annotate %>%
  group_by(gene_id) %>%
  slice_min(`E.value..SwissProt.`, n = 1, with_ties = FALSE) %>%
  ungroup()

# Step 4: Select key columns for the final annotation table: gene_id, protein_name, BLAST description, GO terms, KEGG_ID
aip_annotate <- aip_annotate %>%
  dplyr::select(gene_id, protein_name, Protein_ID,
         starts_with("Hit_description"),
         starts_with("GO_"),
         KEGG_ID)


length(unique(aip_annotate$gene_id))

#Step 5: Save annotation file 
write.csv(aip_annotate, file = file.path(path.out,"aip_annotate.csv"))
```

## 7.2 Parse Bacterial Annotations from GFF3

The three bacterial genomes were annotated with [Bakta](https://github.com/oschwengers/bakta). The GFF3 files contain rich `Name=` (product), `gene=`, `KEGG:`, and `COG:` fields in the attributes column. We parse CDS features to build annotation tables matching the count matrix gene IDs (`lab_AODPDA_XXXXX`, `rue_LDIBIJ_XXXXX`, `vib_BPEJFA_XXXXX`).

```{r parse-bacterial-gff3}
parse_bakta_gff3 <- function(gff3_path, prefix) {
  lines <- readLines(gff3_path)
  lines <- lines[!grepl("^#", lines) & nchar(lines) > 0]

  gff <- read.delim(textConnection(lines), header = FALSE,
                    col.names = c("seqid","source","type","start","end",
                                  "score","strand","phase","attributes"))

  gff_cds <- gff %>% filter(type == "CDS")

  gff_cds <- gff_cds %>%
    mutate(
      locus_tag  = str_extract(attributes, "(?<=locus_tag=)[^;]+"),
      gene_name  = str_extract(attributes, "(?<=;gene=|^gene=)[^;]+"),
      product    = str_extract(attributes, "(?<=Name=)[^;]+") %>% URLdecode(),
      kegg       = str_extract(attributes, "(?<=KEGG:)[^,;]+"),
      cog_id     = str_extract(attributes, "(?<=COG:COG)\\d+") %>%
                     {ifelse(is.na(.), NA, paste0("COG", .))},
      go_terms   = str_extract(attributes, "(?<=Dbxref=).*") %>%
                     str_extract_all("GO:\\d+") %>%
                     sapply(function(x) paste(x, collapse = ";"))
    ) %>%
    mutate(
      gene_id = paste0(prefix, "_", locus_tag),
      go_terms = ifelse(go_terms == "", NA, go_terms)
    ) %>%
    dplyr::select(gene_id, locus_tag, gene_name, product, kegg, cog_id, go_terms) %>%
    distinct(gene_id, .keep_all = TRUE)

  return(gff_cds)
}

annot_lab <- parse_bakta_gff3(file.path(path.in,"labrenzia.gff3"), "lab")
annot_rue <- parse_bakta_gff3(file.path(path.in,"ruegeria.gff3"),  "rue")
annot_vib <- parse_bakta_gff3(file.path(path.in,"vibrio.gff3"), "vib")

annot_bact <- bind_rows(
  annot_lab %>% mutate(organism = "Labrenzia"),
  annot_rue %>% mutate(organism = "Ruegeria"),
  annot_vib %>% mutate(organism = "Vibrio")
)

cat("Labrenzia:", nrow(annot_lab), "CDS,", sum(!is.na(annot_lab$gene_name)), "with gene names,",
    sum(!is.na(annot_lab$kegg)), "with KEGG\n")
cat("Ruegeria:", nrow(annot_rue), "CDS,", sum(!is.na(annot_rue$gene_name)), "with gene names,",
    sum(!is.na(annot_rue$kegg)), "with KEGG\n")
cat("Vibrio:", nrow(annot_vib), "CDS,", sum(!is.na(annot_vib$gene_name)), "with gene names,",
    sum(!is.na(annot_vib$kegg)), "with KEGG\n")

#save annotation file
write.csv(annot_bact, file = file.path(path.out,'annot_bact.csv'))
```

## 7.3 Parse Symbiont Annotations from Combined GTF

The *Symbiodiniaceae* genes in the combined GTF have `gene` and `product` attributes on transcript lines. We extract one annotation per gene_id.

```{r parse-symbiont-gtf}
gtf_lines <- readLines(file.path(path.in,"combined_annotation.gtf"))
sym_transcript_lines <- gtf_lines[grepl("^sym_", gtf_lines) & grepl("\\ttranscript\\t", gtf_lines)]

annot_sym <- data.frame(raw = sym_transcript_lines, stringsAsFactors = FALSE) %>%
  mutate(
    gene_id   = str_extract(raw, 'gene_id "([^"]+)"', group = 1),
    gene_name = str_extract(raw, 'gene "([^"]+)"', group = 1),
    product   = str_extract(raw, 'product "([^"]+)"', group = 1)
  ) %>%
  dplyr::select(-raw) %>%
  filter(!is.na(gene_id)) %>%
  distinct(gene_id, .keep_all = TRUE)

# Create a label column: prefer gene_name, then non-hypothetical product, then gene_id
annot_sym <- annot_sym %>%
  mutate(label = case_when(
    !is.na(gene_name) ~ gene_name,
    !is.na(product) & product != "hypothetical protein" ~ str_trunc(product, 40),
    TRUE ~ gene_id
  ))

cat("Symbiont:", nrow(annot_sym), "genes,",
    sum(!is.na(annot_sym$gene_name)), "with gene names,",
    sum(!is.na(annot_sym$product) & annot_sym$product != "hypothetical protein"),
    "with non-hypothetical products\n")

#save annotation file 
write.csv(annot_sym, file = file.path(path.out,'annot_sym.csv'))
```

# 8. Save data

```{r}
save(list=c('aip_annotate', 'counts_aip', 'colData'),file = file.path(path.out,'host_data.RData'))
```

```{r session-info}
sessionInfo()
```

