---
title: "Metatranscriptomic Analysis of Bacterial Inoculation in Aiptasia"
author: "Hailey Hatch & Maggie Sogin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
---

# Introduction

This notebook analyzes metatranscriptomic (RNA-seq) data from an *Aiptasia* bacterial inoculation experiment. *Aiptasia* were first treated with antibiotics to reduce their native microbiome, then inoculated with one of three bacterial symbionts to examine changes in host, algal symbiont, and bacterial gene expression.


**Treatments:**

- **control/FASW** — no antibiotics, no inoculation (natural microbiome)
- **ABS** — antibiotics only (microbiome-depleted baseline)
- **ISOA/Labrenzia** — antibiotics + *Labrenzia aggregata* inoculation
- **ISOB/Ruegeria** — antibiotics + *Ruegeria mobilis* inoculation
- **ISOC/Vibrio** — antibiotics + *Vibrio alginolyticus* inoculation
- **ISOABC/multi** — antibiotics + all three bacteria combined

**Timepoints:** 1 day post-inoculation (1dpi) and 21 days post-inoculation (21dpi)

**Replicates:** 4 biological replicates per treatment per timepoint = **48 samples total**

**Previous pipeline:** Raw reads were processed with fastp (quality trimming) → SortMeRNA (rRNA removal) → HISAT2 (alignment to a combined reference genome containing *Aiptasia* host, *Symbiodiniaceae* symbiont, *Labrenzia*, *Ruegeria*, and *Vibrio*) → featureCounts (gene-level quantification).

**Biological questions:**

1. What are the general expression patterns across treatments and timepoints?
2. How does bacterial inoculation affect host immune gene expression?
3. Are there changes in sulfur cycling genes, particularly in *Labrenzia* and *Ruegeria*?

**Updated questions**
1. Is there a common host response to bacteria inoculation regardless of treatment?
- compare ABS to each treatment at each timepoint 

2. How does host expression change when inoculated with vibrio alone vs. in multi-species combinations? 
Hypotheses: 
- Host gene expression is more similar with inoculations of beneficial microbes vs pathogen
- Inoculation of multispecies together causes differences in host gene expression 
- Inoculation of pathogen results in DE expression of key genes involved in immune systems and the presence of other bacteria dampens that response (idea --> BMCs do not want to be targeted?)

3. Are there changes in host gene expression that work to recognizing BMCs and not pathogens? 


```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,           # Show code in the output
  warning = FALSE,       # Hide warning messages
  message = FALSE,       # Hide package loading messages
  fig.width = 10,        # Default figure width (inches)
  fig.height = 6,        # Default figure height (inches)
  fig.align = "center"   # Center all figures
)
```

---

# 1. Load Libraries & Define Constants

```{r load-libraries}
library(readxl)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(matrixStats)
library(ggrepel)
library(patchwork)
library(UpSetR)
library(stringr)
```

```{r define-constants}
palette <- c(
  "FASW" = 'grey60',
  "ABS"     = 'black',
  "Labrenzia"    = '#A6CEE3',
  "Ruegeria"    = '#1F78B4',
  "Vibrio"    = '#B2DF8A',
  "multi"  = 'violet')
treatment_levels <- c(
  "control" = "FASW", 
  "ABS" = "ABS", 
  "ISOA" = "Labrenzia", 
  "ISOB" = "Ruegeria", 
  "ISOC" = "Vibrio", 
  "ISOABC" = "multi")

path.out<-"../../output/transcriptomics"
path.in<-"../../data/transcriptomics"
```

---

# 2. Load All Data Files

```{r load-all-data}
# featureCounts gene count matrix (6 annotation cols + 48 samples)
counts_raw <- read.delim(file.path(path.in,"counts.tsv"), comment.char = "#", check.names = FALSE)

# featureCounts assignment summary
counts_summary <- read.delim(file.path(path.in,"counts.tsv.summary"), check.names = FALSE)

# HISAT2 alignment rates
hisat2_sum <- read.delim(file.path(path.in,"hisat2_alignment_summary.tsv"))

# Per-reference organism read counts (aip, sym, lab, rue, vib)
organism_counts <- read.delim(file.path(path.in,"per_reference_counts.tsv"))

# SortMeRNA rRNA removal summary
sortme_sum <- read.delim(file.path(path.in,"sortmerna_summary.tsv"))

# Fastp trimming summary
fastp_sum <- read.delim(file.path(path.in,"fastp_summary_per_sample.tsv"), check.names = FALSE)

# Aiptasia gene annotations (SwissProt/TrEMBL BLAST, GO, KEGG)
aip_annotate <- read.delim(file.path(path.in,"Aiptasia_annotation_file.csv"))
```

---

# 3. Build Sample Metadata

Create the `colData` data frame that DESeq2 requires. We'll parse treatment and timepoint directly from the sample names so it's fully reproducible.

```{r build-metadata}
# Step 1: Make function to clean sample names
clean_sample_name <- function(x) {gsub("_S\\d+.*$", "", x)}

# Step 2: Get sample column names from the counts data
colnames(counts_raw) <- clean_sample_name(basename(colnames(counts_raw)))
annotation_cols <- c("Geneid", "Chr", "Start", "End", "Strand", "Length")
sample_names <- colnames(counts_raw)[!colnames(counts_raw) %in% annotation_cols]

# Step 3: Parse treatment and timepoint from the clean names
treatment <- str_extract(sample_names, "ISOABC|ISOA|ISOB|ISOC|ABS|control")
timepoint <- str_extract(sample_names, "\\d+dpi")

# Step 4: Create the colData data frame
  colData <- data.frame(row.names = sample_names, treatment = treatment, timepoint = timepoint)

# Step 5: Convert treatment and timepoint to factors with your defined levels
colData$treatment <- factor(treatment_levels[as.character(colData$treatment)], 
                            levels = treatment_levels)
colData$timepoint <- factor(colData$timepoint, levels = c("1dpi", "21dpi"))

stopifnot(all(table(colData$treatment, colData$timepoint) == 4))

# Make metadata by turning rownames into a regular column
metadata <- colData %>% rownames_to_column("sample")
```

---

# 4. QC Summary Plots

Visualize pipeline quality metrics to identify any problematic samples before DESeq2.

## 4.1 Fastp — Read Quality & Trimming

```{r plot-fastp}
plot_QC_fastp <- fastp_sum %>%
  mutate(sample = clean_sample_name(sample)) %>%
  left_join(metadata, by = "sample") %>%
  filter(!is.na(treatment)) %>%
  arrange(treatment, timepoint) %>%
  mutate(sample = fct_inorder(sample)) %>%
  dplyr::select(sample, treatment, timepoint, reads_before, reads_after) %>%
  pivot_longer(cols = c(reads_before, reads_after),
               names_to = "stage", values_to = "reads") %>%
  mutate(stage = factor(stage, levels = c("reads_before", "reads_after"),
                        labels = c("Raw", "After Trimming"))) %>%
  ggplot(aes(x = sample, y = reads / 1e6, fill = stage)) +
  geom_col(position = "dodge") +
  facet_wrap(~ timepoint, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Fastp: Read Counts Before & After Trimming",
       y = "Reads (millions)", x = NULL)
plot_QC_fastp
# No extreme trimming; all samples ~125-225M reads
```

## 4.2 SortMeRNA — rRNA Removal Rates

```{r plot-sortmerna}
plot_QC_sortmerna <- sortme_sum %>%
  mutate(sample = clean_sample_name(sample)) %>%
  left_join(metadata, by = "sample") %>%
  arrange(treatment, timepoint) %>%
  mutate(sample = fct_inorder(sample)) %>%
  ggplot(aes(x = sample, y = rRNA_pct, fill = treatment)) +
  geom_col() +
  facet_wrap(~ timepoint, scales = "free_x") +
  scale_fill_manual(values = palette) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "SortMeRNA: rRNA Removal Rates", y = "rRNA (%)", x = NULL)
plot_QC_sortmerna
# ~5-20% rRNA; 21dpi slightly higher on average
```

## 4.3 HISAT2 — Alignment Rates

```{r plot-hisat2}
plot_QC_hisat2 <- hisat2_sum %>%
  mutate(sample = clean_sample_name(sample)) %>%
  left_join(metadata, by = "sample") %>%
  arrange(treatment, timepoint) %>%
  mutate(sample = fct_inorder(sample)) %>%
  ggplot(aes(x = sample, y = `overall_alignment_rate...`, fill = treatment)) +
  geom_col() +
  facet_wrap(~ timepoint, scales = "free_x") +
  scale_fill_manual(values = palette) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "HISAT2: Alignment Rates", y = "Alignment Rate (%)", x = NULL)
plot_QC_hisat2
# Alignment rates 50-75%
```

## 4.4 Supplementary Fig S1 — % Library Mapped to Each Genome

```{r fig-s1-pct-mapped, fig.width=12, fig.height=6}
# Compute % of total mapped reads assigned to each organism per sample
figS1_data <- organism_counts %>%
  mutate(sample = clean_sample_name(sample)) %>%
  left_join(metadata, by = "sample") %>%
  mutate(total = aip + sym + lab + rue + vib) %>%
  pivot_longer(cols = c(aip, sym, lab, rue, vib),
               names_to = "organism", values_to = "reads") %>%
  mutate(pct_mapped = 100 * reads / total,
         organism = factor(organism,
                           levels = c("aip", "sym", "lab", "rue", "vib"),
                           labels = c("Host", "Symbiodinium", "Labrenzia", "Ruegeria", "Vibrio")),
         treatment = factor(treatment, levels = treatment_levels))

plot_figS1 <- ggplot(figS1_data, aes(x = treatment, y = pct_mapped, color = organism)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 2.5, alpha = 0.8) +
  facet_grid(organism ~ timepoint, scales = "free_y") +
  scale_color_manual(values = c("Host" = "#1b9e77", "Symbiodinium" = "#d95f02",
                                "Labrenzia" = "#7570b3", "Ruegeria" = "#e7298a",
                                "Vibrio" = "#66a61e")) +
  ylim(0, NA) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        strip.text = element_text(face = "bold")) +
  labs(title = "Fig S1 — % Library Mapped to Each Genome",
       y = "% of reads mapped to genome", x = NULL)
plot_figS1

ggsave(file.path(path.out,"FigS1_pct_mapped_by_genome.pdf"), plot_figS1, width = 12, height = 6)
ggsave(file.path(path.out,"FigS1_pct_mapped_by_genome.png"), plot_figS1, width = 12, height = 6, dpi = 300)
```

## 4.5 featureCounts — Read Assignment Summary

```{r plot-featurecounts}
colnames(counts_summary) <- clean_sample_name(basename(colnames(counts_summary)))

fc_data <- counts_summary %>%
  pivot_longer(cols = -Status, names_to = "sample", values_to = "reads") %>%
  filter(reads > 0) %>%
  left_join(metadata, by = "sample") %>%
  arrange(treatment, timepoint) %>%
  mutate(sample = fct_inorder(sample))

plot_QC_featurecounts <- ggplot(fc_data, aes(x = sample, y = reads, fill = Status)) +
  geom_col(position = "fill") +
  facet_wrap(~ timepoint, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "featureCounts: Read Assignment", y = "Proportion", x = NULL)
plot_QC_featurecounts
# Assigned reads ~25% — low but expected for metatranscriptomic reference
```

## 4.6 featureCounts — Library Size

```{r plot-library-size}
plot_QC_library_size <- fc_data %>%
  filter(Status == "Assigned") %>%
  ggplot(aes(x = sample, y = reads / 1e6, fill = treatment)) +
  geom_col() +
  facet_wrap(~ timepoint, scales = "free_x") +
  scale_fill_manual(values = palette) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Library Size (Assigned Reads)", y = "Reads (millions)", x = NULL)
plot_QC_library_size
# ~20-40M reads assigned to combined reference per sample
```

## 4.7 Save QC Plots

```{r save-qc-plots}
qc_plots <- list(
  "plot_QC_fastp"         = plot_QC_fastp,
  "plot_QC_sortmerna"     = plot_QC_sortmerna,
  "plot_QC_hisat2"        = plot_QC_hisat2,
  "plot_QC_featurecounts" = plot_QC_featurecounts,
  "plot_QC_library_size"  = plot_QC_library_size)

for (name in names(qc_plots)) {
  ggsave(paste0(path.out, name, ".pdf"), qc_plots[[name]], width = 10, height = 6)
  ggsave(paste0(path.out, name, ".png"), qc_plots[[name]], width = 10, height = 6, dpi = 300)
}
```

---

# 5. Clean and Prepare the Count Matrix

Transform the raw featureCounts output into a clean integer count matrix that DESeq2 can use.

```{r clean-counts}
# Save gene lengths for TPM calculation later
gene_lengths <- counts_raw %>% 
                dplyr::select(Geneid, Length) %>% column_to_rownames("Geneid")

counts_matrix <- counts_raw %>%
  dplyr::select(-Chr, -Start, -End, -Strand, -Length) %>%
  column_to_rownames("Geneid") %>%
  as.matrix()
counts_matrix <- round(counts_matrix)

all(colnames(counts_matrix) == rownames(colData))

dim(counts_matrix)
head(counts_matrix)
```

## 5.1 Compute TPM Matrix

TPM (Transcripts Per Million) normalizes for both gene length and sequencing depth,
making expression values comparable across samples AND genes

TPM formula: for each gene in each sample,
1. Divide raw count by gene length in kb → RPK (reads per kilobase)
2. Sum all RPKs in that sample → scaling factor
3. Divide each RPK by (scaling factor / 1e6) → TPM

```{r compute-tpm}
all(rownames(gene_lengths) == rownames(counts_matrix))

gene_lengths_kb <- gene_lengths$Length / 1000

rpk_matrix <- sweep(counts_matrix, 1, gene_lengths_kb, "/")

scaling_factors <- colSums(rpk_matrix)

tpm_matrix <- sweep(rpk_matrix, 2, scaling_factors / 1e6, "/")

tpm_aip <- tpm_matrix[startsWith(rownames(tpm_matrix), "aip_"), ]
tpm_sym <- tpm_matrix[startsWith(rownames(tpm_matrix), "sym_"), ]
tpm_lab <- tpm_matrix[startsWith(rownames(tpm_matrix), "lab_"), ]
tpm_rue <- tpm_matrix[startsWith(rownames(tpm_matrix), "rue_"), ]
tpm_vib <- tpm_matrix[startsWith(rownames(tpm_matrix), "vib_"), ]

```

---

# 6. Split Counts by Organism

Analyzing organisms separately because:
- Library size normalization should be computed within an organism
- The biological questions differ per organism
- Read depths vary (host: millions, bacteria: thousands)

```{r split-by-organism}
counts_aip <- counts_matrix[startsWith(rownames(counts_matrix), "aip_"), ]
counts_sym <- counts_matrix[startsWith(rownames(counts_matrix), "sym_"), ]
counts_lab <- counts_matrix[startsWith(rownames(counts_matrix), "lab_"), ]
counts_rue <- counts_matrix[startsWith(rownames(counts_matrix), "rue_"), ]
counts_vib <- counts_matrix[startsWith(rownames(counts_matrix), "vib_"), ]

# Verify splits include all counts
nrow(counts_aip) + nrow(counts_sym) + nrow(counts_lab) + nrow(counts_rue) + nrow(counts_vib) == nrow(counts_matrix)
```

Gene filtering
- Below calculate the Nr of genes passing the filter, a gene is included in the filter if it has at least 10 reads in 4 samples across the entire dataset (Ref: deseq manual )

```{r organism-summary-stats}
organism_list <- list(
  aip = counts_aip, sym = counts_sym,
  lab = counts_lab, rue = counts_rue, vib = counts_vib)

organism_summary <- data.frame(
  organism = names(organism_list),
  total_genes = sapply(organism_list, nrow),
  median_counts_per_sample = sapply(organism_list, function(x) median(colSums(x))), #median nr. reads mapping to each sample
  genes_all_zero = sapply(organism_list, function(x) sum(rowSums(x) == 0)), #nr genes no expression data at all 
  genes_passing_filter = sapply(organism_list, function(x) sum(rowSums(x >= 10) >= 4)) # at least 10 counts in 4 samps
)
organism_summary
# Few bacterial genes passed the filter
```

```{r save-organism-counts}
for (name in names(organism_list)) {
  write.table(organism_list[[name]],
              file = paste0(path.out,"counts_", name, ".tsv"),
              sep = "\t", quote = FALSE, col.names = NA)}
```

---

# 7. Prepare Annotation Data

The annotation file has multiple transcript isoforms per gene (e.g., g1.t1 and g1.t2 are isoforms of gene g1). We need to:

1. Extract the gene-level ID
2. Match it to the count matrix format (with `aip_` prefix)
3. Deduplicate to one annotation per gene 

## 7.1 Parse Aiptasia annotations
```{r prepare-annotations}
# Step 1: Extract gene ID from Protein_ID
aip_annotate <- aip_annotate %>%
  mutate(gene_id = paste0("aip_", sub("\\..*$", "", Protein_ID)))

# Step 2: Extract a short protein name from the SwissProt BLAST hit column
aip_annotate <- aip_annotate %>%
  mutate(protein_name = str_extract(`BLAST_hit..SwissProt.`, "(?<=\\|)[^|]+(?=_)"))

# Step 3: Deduplicate — keep one row per gene_id
aip_annotate <- aip_annotate %>%
  group_by(gene_id) %>%
  slice_min(`E.value..SwissProt.`, n = 1, with_ties = FALSE, na_rm = TRUE) %>%
  ungroup()

# Step 4: Select key columns for the final annotation table: gene_id, protein_name, BLAST description, GO terms, KEGG_ID
aip_annotate <- aip_annotate %>%
  dplyr::select(gene_id, protein_name, Protein_ID,
         starts_with("Hit_description"),
         starts_with("GO_"),
         KEGG_ID)
```

## 7.2 Parse Bacterial Annotations from GFF3

The three bacterial genomes were annotated with [Bakta](https://github.com/oschwengers/bakta). The GFF3 files contain rich `Name=` (product), `gene=`, `KEGG:`, and `COG:` fields in the attributes column. We parse CDS features to build annotation tables matching the count matrix gene IDs (`lab_AODPDA_XXXXX`, `rue_LDIBIJ_XXXXX`, `vib_BPEJFA_XXXXX`).

```{r parse-bacterial-gff3}
library(dplyr)
parse_bakta_gff3 <- function(gff3_path, prefix) {
  lines <- readLines(gff3_path)
  lines <- lines[!grepl("^#", lines) & nchar(lines) > 0]

  gff <- read.delim(textConnection(lines), header = FALSE,
                    col.names = c("seqid","source","type","start","end",
                                  "score","strand","phase","attributes"))

  gff_cds <- gff %>% filter(type == "CDS")

  gff_cds <- gff_cds %>%
    mutate(
      locus_tag  = str_extract(attributes, "(?<=locus_tag=)[^;]+"),
      gene_name  = str_extract(attributes, "(?<=;gene=|^gene=)[^;]+"),
      product    = str_extract(attributes, "(?<=Name=)[^;]+") %>% URLdecode(),
      kegg       = str_extract(attributes, "(?<=KEGG:)[^,;]+"),
      cog_id     = str_extract(attributes, "(?<=COG:COG)\\d+") %>%
                     {ifelse(is.na(.), NA, paste0("COG", .))},
      go_terms   = str_extract(attributes, "(?<=Dbxref=).*") %>%
                     str_extract_all("GO:\\d+") %>%
                     sapply(function(x) paste(x, collapse = ";"))
    ) %>%
    mutate(
      gene_id = paste0(prefix, "_", locus_tag),
      go_terms = ifelse(go_terms == "", NA, go_terms)
    ) %>%
    dplyr::select(gene_id, locus_tag, gene_name, product, kegg, cog_id, go_terms) %>%
    distinct(gene_id, .keep_all = TRUE)

  return(gff_cds)
}

annot_lab <- parse_bakta_gff3(file.path(path.in,"labrenzia.gff3"), "lab")
annot_rue <- parse_bakta_gff3(file.path(path.in,"ruegeria.gff3"),  "rue")
annot_vib <- parse_bakta_gff3(file.path(path.in,"vibrio.gff3"), "vib")

annot_bact <- bind_rows(
  annot_lab %>% mutate(organism = "Labrenzia"),
  annot_rue %>% mutate(organism = "Ruegeria"),
  annot_vib %>% mutate(organism = "Vibrio")
)

cat("Labrenzia:", nrow(annot_lab), "CDS,", sum(!is.na(annot_lab$gene_name)), "with gene names,",
    sum(!is.na(annot_lab$kegg)), "with KEGG\n")
cat("Ruegeria:", nrow(annot_rue), "CDS,", sum(!is.na(annot_rue$gene_name)), "with gene names,",
    sum(!is.na(annot_rue$kegg)), "with KEGG\n")
cat("Vibrio:", nrow(annot_vib), "CDS,", sum(!is.na(annot_vib$gene_name)), "with gene names,",
    sum(!is.na(annot_vib$kegg)), "with KEGG\n")
```

## 7.3 Parse Symbiont Annotations from Combined GTF

The *Symbiodiniaceae* genes in the combined GTF have `gene` and `product` attributes on transcript lines. We extract one annotation per gene_id.

```{r parse-symbiont-gtf}
gtf_lines <- readLines(file.path(path.in,"combined_annotation.gtf"))
sym_transcript_lines <- gtf_lines[grepl("^sym_", gtf_lines) & grepl("\\ttranscript\\t", gtf_lines)]

annot_sym <- data.frame(raw = sym_transcript_lines, stringsAsFactors = FALSE) %>%
  mutate(
    gene_id   = str_extract(raw, 'gene_id "([^"]+)"', group = 1),
    gene_name = str_extract(raw, 'gene "([^"]+)"', group = 1),
    product   = str_extract(raw, 'product "([^"]+)"', group = 1)
  ) %>%
  dplyr::select(-raw) %>%
  filter(!is.na(gene_id)) %>%
  distinct(gene_id, .keep_all = TRUE)

# Create a label column: prefer gene_name, then non-hypothetical product, then gene_id
annot_sym <- annot_sym %>%
  mutate(label = case_when(
    !is.na(gene_name) ~ gene_name,
    !is.na(product) & product != "hypothetical protein" ~ str_trunc(product, 40),
    TRUE ~ gene_id
  ))

cat("Symbiont:", nrow(annot_sym), "genes,",
    sum(!is.na(annot_sym$gene_name)), "with gene names,",
    sum(!is.na(annot_sym$product) & annot_sym$product != "hypothetical protein"),
    "with non-hypothetical products\n")
```

---

# 8. DESeq2 Analysis — Host (*Aiptasia*)

Host differential expression analysis using DESeq2. DESeq2 models RNA-seq count data using a negative binomial distribution, accounts for library size differences between samples, and uses shrinkage to produce more reliable fold-change estimates.

The design should include three terms: timepoint, teratment and treatment:timepoint interaction

Note (June 29): there is an alternatively approach that we could try, namely aggragete gene counts within samples by GO categegory and do a DEG analysis of those counts, the resolution will be limited but could also be more ecologically informative? Need to think on this as we are already getting good seperate below

## 8.1 Create DESeq object, filter data and run LRT test

https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#countmat
As per recommendations in Interactions section: 

```{r specify models and generate DESeq Object}
dds_aip <- DESeqDataSetFromMatrix(
  countData = counts_aip,
  colData = colData,
  design = ~ timepoint*treatment)

cat(nrow(dds_aip), "genes before filtering\n")
keep <- rowSums(counts(dds_aip) >= 10) >= 4
dds_aip <- dds_aip[keep, ]
cat(nrow(dds_aip), "genes after filtering\n")
#removes genes that don't have at least 10 counts in at least 4 samples, this helps to reduced noise in analysis 
```

## 8.2 Run DESeq analysis 
want to look at differences in gene expression at each time-point across treatment groups
```{r}
ds<-DESeq(dds_aip)
plotDispEsts(ds)
# Dispersion plot looks good: trend line decreases smoothly, gene-wise estimates converge (black dots)
```


## 8.3 Exploratory plots

### VST Transformation

```{r vst-aip}
vsd_aip <- vst(dds_aip, blind = TRUE)
```

### PCA — All Samples
```{r pca-aip-all}
pca_data <- plotPCA(vsd_aip, intgroup = c("treatment", "timepoint"), returnData = TRUE)
percent_var <- round(100 * attr(pca_data, "percentVar"))

p <- ggplot(pca_data, aes(x = PC1, y = PC2, color = treatment, shape = timepoint)) +
  geom_point(size = 3) +
  scale_color_manual(values = palette) +
  theme_bw() +
  labs(x = paste0("PC1: ", percent_var[1], "% variance"),
       y = paste0("PC2: ", percent_var[2], "% variance"),
       title = "PCA — Aiptasia")
p
ggsave(file.path(path.out,"pca_host_all.pdf"), p, width = 7, height = 5)
# Timepoint is the dominant source of variation, although controls at 1dpi cluster with all 21dpi, suggesting variation largely due to ABS influence
```
Impacts of the antibiotics is negligble at day 21

```{r PERMANOVA}
library(vegan)
# get VST matrix (samples as rows)
vsd_mat <- t(assay(vsd_aip))

# pull metadata
meta <- as.data.frame(colData(vsd_aip))

# run PERMANOVA
perm_res <- adonis2(vsd_mat ~ treatment+timepoint,
                    data = meta,
                    method = "euclidean",
                    permutations = 999,
                    by="terms")
perm_res
```
Significant effect of time but not treatment on expression profiles
Suggests that expression differences are related to ABS exposure and not to microbial inoculations. 


### Sampel to Sample distance matrix
```{r}
library("RColorBrewer")
sampleDists <- dist(vsd_mat)
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd_aip$timepoint, vsd_aip$treatment, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```


## 8.4 Contrasts
### 1dpi
```{r 1 dpi contrasts}
tp<-'1dpi'
dds_aip_1dpi <- ds[, ds$timepoint == tp]
dds_aip_1dpi$timepoint <- droplevels(dds_aip_1dpi$timepoint)
design(dds_aip_1dpi) <- ~ treatment
dds_aip_1dpi <- DESeq(dds_aip_1dpi)

# contrasts of all treatments in comparison to ABS
res_aip_1dpi_control <- results(dds_aip_1dpi, contrast = c("treatment", "FASW", "ABS"))
res_aip_1dpi_ISOA <- results(dds_aip_1dpi, contrast = c("treatment", "Labrenzia", "ABS"))
res_aip_1dpi_ISOB <- results(dds_aip_1dpi, contrast = c("treatment", "Ruegeria", "ABS"))
res_aip_1dpi_ISOC <- results(dds_aip_1dpi, contrast = c("treatment", "Vibrio", "ABS"))
res_aip_1dpi_ISOABC <- results(dds_aip_1dpi, contrast = c("treatment", "multi", "ABS"))

# multi infection contrasts
res_aip_1dpi_ISOA_vs_ISOABC <- results(dds_aip_1dpi, contrast = c("treatment", "Labrenzia", "multi"))
res_aip_1dpi_ISOB_vs_ISOABC <- results(dds_aip_1dpi, contrast = c("treatment", "Ruegeria", "multi"))
res_aip_1dpi_ISOC_vs_ISOABC <- results(dds_aip_1dpi, contrast = c("treatment", "Vibrio", "multi"))
```

### 21 dpi
```{r 1 dpi contrasts}
tp<-'21dpi'
dds_aip_21dpi <- ds[, ds$timepoint == tp]
dds_aip_21dpi$timepoint <- droplevels(dds_aip_21dpi$timepoint)
design(dds_aip_21dpi) <- ~ treatment
dds_aip_21dpi <- DESeq(dds_aip_21dpi)

# contrasts of all treatments in comparison to ABS
res_aip_21dpi_control <- results(dds_aip_21dpi, contrast = c("treatment", "FASW", "ABS"))
res_aip_21dpi_ISOA <- results(dds_aip_21dpi, contrast = c("treatment", "Labrenzia", "ABS"))
res_aip_21dpi_ISOB <- results(dds_aip_21dpi, contrast = c("treatment", "Ruegeria", "ABS"))
res_aip_21dpi_ISOC <- results(dds_aip_21dpi, contrast = c("treatment", "Vibrio", "ABS"))
res_aip_21dpi_ISOABC <- results(dds_aip_21dpi, contrast = c("treatment", "multi", "ABS"))

# multi infection contrasts
res_aip_21dpi_ISOA_vs_ISOABC <- results(dds_aip_21dpi, contrast = c("treatment", "Labrenzia", "multi"))
res_aip_21dpi_ISOB_vs_ISOABC <- results(dds_aip_21dpi, contrast = c("treatment", "Ruegeria", "multi"))
res_aip_21dpi_ISOC_vs_ISOABC <- results(dds_aip_21dpi, contrast = c("treatment", "Vibrio", "multi"))
```

## 8.5 Annotation tables of DEGs
GO IDs are swissprot annotations
```{r - define function to make table}
make_deg_table <- function(comp_data, comparison, timepoint, annotation_table = aip_annotate) {
  res <- data.frame(comp_data) %>%
    filter(padj < 0.05) %>% #change if needed
    rownames_to_column(var = "gene_id") %>%
    left_join(annotation_table, by = "gene_id") %>%
    mutate(comp = comparison, timepoint=timepoint) %>%
    mutate(
      'go_ids'=GO_IDs..SwissProt., 
      'go_cat'=GO_categories..SwissProt., 
      'go_bp'=GO_BP..SwissProt.,
      'go_mf'=GO_MF..SwissProt.,
      'go_cc'=GO_CC..SwissProt.) %>% 
    dplyr::select(
      gene_id, comp,timepoint, baseMean, log2FoldChange, lfcSE, stat, padj,
      protein_name, go_ids, go_cat,
      go_bp, go_mf, go_cc, KEGG_ID
    )
  return(res)
}
```

###1 dpi
```{r - apply to each deg set}
deg_set_fasw_v_abs_1dpi<-make_deg_table(res_aip_1dpi_control,'FASW_vs_ABS','1dpi',aip_annotate)
deg_set_lab_v_abs_1dpi<-make_deg_table(res_aip_1dpi_ISOA,'Labrenzia_vs_ABS','1dpi',aip_annotate)
deg_set_reu_v_abs_1dpi<-make_deg_table(res_aip_1dpi_ISOB,'Reugeria_vs_ABS','1dpi',aip_annotate)
deg_set_vib_v_abs_1dpi<-make_deg_table(res_aip_1dpi_ISOC,'Vibrio_vs_ABS','1dpi',aip_annotate)
deg_set_multi_v_abs_1dpi<-make_deg_table(res_aip_1dpi_ISOABC,'Multi_vs_ABS','1dpi',aip_annotate)

deg_set_lab_v_multi_1dpi<-make_deg_table(res_aip_1dpi_ISOA_vs_ISOABC,'ISOA_vs_ISOABC','1dpi',aip_annotate)
deg_set_reu_v_multi_1dpi<-make_deg_table(res_aip_1dpi_ISOB_vs_ISOABC,'ISOB_vs_ISOABC','1dpi',aip_annotate)
deg_set_vib_v_multi_1dpi<-make_deg_table(res_aip_1dpi_ISOC_vs_ISOABC,'ISOC_vs_ISOABC','1dpi',aip_annotate)
```
### 21dpi
```{r - apply to each deg set}
deg_set_fasw_v_abs_21dpi<-make_deg_table(res_aip_21dpi_control,'FASW_vs_ABS','21dpi',aip_annotate)
deg_set_lab_v_abs_21dpi<-make_deg_table(res_aip_21dpi_ISOA,'Labrenzia_vs_ABS','21dpi',aip_annotate)
deg_set_reu_v_abs_21dpi<-make_deg_table(res_aip_21dpi_ISOB,'Reugeria_vs_ABS','21dpi',aip_annotate)
deg_set_vib_v_abs_21dpi<-make_deg_table(res_aip_21dpi_ISOC,'Vibrio_vs_ABS','21dpi',aip_annotate)
deg_set_multi_v_abs_21dpi<-make_deg_table(res_aip_21dpi_ISOABC,'Multi_vs_ABS','21dpi',aip_annotate)

deg_set_lab_v_multi_21dpi<-make_deg_table(res_aip_21dpi_ISOA_vs_ISOABC,'ISOA_vs_ISOABC','21dpi',aip_annotate)
deg_set_reu_v_multi_21dpi<-make_deg_table(res_aip_21dpi_ISOB_vs_ISOABC,'ISOB_vs_ISOABC','21dpi',aip_annotate)
deg_set_vib_v_multi_21dpi<-make_deg_table(res_aip_21dpi_ISOC_vs_ISOABC,'ISOC_vs_ISOABC','21dpi',aip_annotate)
```


```{r - combine deg tables}
deg_names <- ls(pattern = "^deg_set")
deg_list <- mget(deg_names) #put dataframes into list 
deg_list <- deg_list[!sapply(deg_list, is.null)]# Drop any NULLs
deg_list <- deg_list[sapply(deg_list, nrow) > 0] # Drop any with 0 rows
combined_deg <- bind_rows(deg_list, .id = "source")
write.csv(combined_deg,file.path(path.out,'DEGs.csv'))
```

## 8.6 Heatmap

- Plot DEGs from each treatment (Labrenzia, Reugeria, Vibrio and Mult) vs ABS comparison 
- Filter genes with padj < 0.05

#### 1 dpi
```{r}
timepoint_label<-'1dpi'
vsd<-vsd_aip

sig_genes <- unique(c(
  rownames(subset(as.data.frame(res_aip_1dpi_ISOA), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOB), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOC), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOABC), padj < 0.05))
))

#get mat and transform to z-score
mat<-assay(vsd[, vsd$timepoint == timepoint_label])[sig_genes, ] 
mat_z<-t(scale(t(mat)))
mat<-mat[complete.cases(mat_z),]

title<-cat(timepoint_label, "union DEGs:", length(sig_genes), "->", nrow(mat_z), "plotted\n")

# Column annotation
ann_col <- data.frame(
    Treatment = vsd[, vsd$timepoint == timepoint_label]$treatment,
    row.names = colnames(mat_z))

#row annotations
row_lab <- aip_annotate$protein_name[match(rownames(mat_z), aip_annotate$gene_id)]

p <- pheatmap(mat_z,
    annotation_col = ann_col,
    annotation_colors = list(
    Treatment = palette),
    show_rownames = TRUE,
    show_colnames = F,
    labels_row = row_lab,
    fontsize_row = 5.5,
    cutree_rows = 3,
    cutree_cols = 4,
    main = paste0("DEGs, Aiptasia ", timepoint_label,
                  " (", nrow(mat_z), " genes), Z-score"),
    silent = TRUE)
grid::grid.newpage(); grid::grid.draw(p$gtable)
```
#### 21 dpi
```{r}
timepoint_label<-'21dpi'
vsd<-vsd_aip

sig_genes <- unique(c(
  rownames(subset(as.data.frame(res_aip_21dpi_ISOA), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOB), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOC), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOABC), padj < 0.05))
))

#get mat and transform to z-score
mat<-assay(vsd[, vsd$timepoint == timepoint_label])[sig_genes, ] 
mat_z<-t(scale(t(mat)))
mat<-mat[complete.cases(mat_z),]

title<-cat(timepoint_label, "union DEGs:", length(sig_genes), "->", nrow(mat_z), "plotted\n")

# Column annotation
ann_col <- data.frame(
    Treatment = vsd[, vsd$timepoint == timepoint_label]$treatment,
    row.names = colnames(mat_z))

#row annotations
row_lab <- aip_annotate$protein_name[match(rownames(mat_z), aip_annotate$gene_id)]

p <- pheatmap(mat_z,
    annotation_col = ann_col,
    annotation_colors = list(
    Treatment = palette),
    show_rownames = TRUE,
    show_colnames = F,
    labels_row = row_lab,
    fontsize_row = 5.5,
    cutree_rows = 3,
    cutree_cols = 4,
    main = paste0("DEGs, Aiptasia ", timepoint_label,
                  " (", nrow(mat_z), " genes), Z-score"),
    silent = TRUE)
grid::grid.newpage(); grid::grid.draw(p$gtable)
```




# Sandbox code


## 8.5 Annotations and DEG tables
Steps: 
1. Convert annotations to long format 
2. select annotations that have p < 0.05
3. Make new annotation df for each comparison, add pvalue and log2fc for each sig gene
4. Make plot showing logfc (x axis) and GO category Y access and size of bubble p-value

```{r convert aip to long format}
library(GO.db)
go_long <-aip_annotate %>%
  dplyr::select(gene_id, GO_IDs..SwissProt.) %>%
  separate_rows(GO_IDs..SwissProt., sep = ";\\s*") %>%
  dplyr::rename(go_id = GO_IDs..SwissProt.) %>%
  mutate(go_id = trimws(go_id)) %>%
  filter(go_id != "", !is.na(go_id)) %>%
  mutate(
    term     = sapply(go_id, function(id) tryCatch(GOTERM[[id]]@Term,     error = function(e) NA)),
    ontology = sapply(go_id, function(id) tryCatch(GOTERM[[id]]@Ontology, error = function(e) NA))
  )
head(go_long)
```

```{r}
length(unique(go_long$term))
length(unique(go_long$go_id))
```



###1 dpi contrasts
#### FASW vs. ABS
```{r}

#set sig genes
res<-data.frame(res_aip_1dpi_control) %>% 
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id") 

degs<-res %>% pull(gene_id) %>% unique()

#get annotation table
annot.tmp<- go_long %>% 
              filter(gene_id %in% degs) 

# merge
df <- annot.tmp %>%
  left_join(res, by = "gene_id")

# Plot

top_n_genes <- 15

# Step 1: find top N gene_ids by padj
top_genes <- df %>%
  distinct(gene_id, padj) %>%
  slice_min(padj, n = top_n_genes) %>%
  pull(gene_id)

# Step 2: filter original df to just those genes
plot_data <- df %>%
  filter(gene_id %in% top_genes) %>%
  distinct(gene_id, term, ontology, log2FoldChange, padj) %>%
  mutate(gene_id = fct_reorder(gene_id, log2FoldChange))

ggplot(plot_data, aes(x = log2FoldChange, y = term, size = -log10(padj), color = ontology)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(name = "-log10(padj)", range = c(2, 10)) +
  labs(
    x = "log2 Fold Change",
    y = "Gene",
    color = "Ontology"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 5))
```

#### Labrenzia vs. ABS
```{r}
#set sig genes
res<-data.frame(res_aip_1dpi_ISOA) %>% 
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id") 

degs<-res %>% pull(gene_id) %>% unique()

#get annotation table
annot.tmp<- go_long %>% 
              filter(gene_id %in% degs) 

# merge
df <- annot.tmp %>%
  left_join(res, by = "gene_id")

# Plot

top_n_genes <- 15

# Step 1: find top N gene_ids by padj
top_genes <- df %>%
  distinct(gene_id, padj) %>%
  slice_min(padj, n = top_n_genes) %>%
  pull(gene_id)

# Step 2: filter original df to just those genes
plot_data <- df %>%
  filter(gene_id %in% top_genes) %>%
  distinct(gene_id, term, ontology, log2FoldChange, padj) %>%
  mutate(gene_id = fct_reorder(gene_id, log2FoldChange))

ggplot(plot_data, aes(x = log2FoldChange, y = term, size = -log10(padj), color = ontology)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(name = "-log10(padj)", range = c(2, 10)) +
  labs(
    x = "log2 Fold Change",
    y = "Gene",
    color = "Ontology"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 5))
```

#### Reugeria vs. ABS
```{r}
#set sig genes
res<-data.frame(res_aip_1dpi_ISOB) %>% 
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id") 

degs<-res %>% pull(gene_id) %>% unique()

#get annotation table
annot.tmp<- go_long %>% 
              filter(gene_id %in% degs) 

# merge
df <- annot.tmp %>%
  left_join(res, by = "gene_id")

# Plot

top_n_genes <- 15

# Step 1: find top N gene_ids by padj
top_genes <- df %>%
  distinct(gene_id, padj) %>%
  slice_min(padj, n = top_n_genes) %>%
  pull(gene_id)

# Step 2: filter original df to just those genes
plot_data <- df %>%
  filter(gene_id %in% top_genes) %>%
  distinct(gene_id, term, ontology, log2FoldChange, padj) %>%
  mutate(gene_id = fct_reorder(gene_id, log2FoldChange))

ggplot(plot_data, aes(x = log2FoldChange, y = term, size = -log10(padj), color = ontology)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(name = "-log10(padj)", range = c(2, 10)) +
  labs(
    x = "log2 Fold Change",
    y = "Gene",
    color = "Ontology"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 5))
```




- Use GO.db to categorize annotations of DEGs
- Generate one DEG tables containing the following:
  - Gene ID, logFC, adj.p value, comparison, dpi

```{r}
#get DEGs
#get sig genes
sig_genes <- unique(c(
  rownames(subset(as.data.frame(res_aip_1dpi_ISOA), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOB), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOC), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOABC), padj < 0.05))
))
```

parse SwissProt GO Terms 
```{r}
library(GO.db)
go_long <-aip_annotate %>%
  dplyr::select(gene_id, GO_IDs..SwissProt.) %>%
  separate_rows(GO_IDs..SwissProt., sep = ";\\s*") %>%
  dplyr::rename(go_id = GO_IDs..SwissProt.) %>%
  mutate(go_id = trimws(go_id)) %>%
  filter(go_id != "", !is.na(go_id)) %>%
  mutate(
    term     = sapply(go_id, function(id) tryCatch(GOTERM[[id]]@Term,     error = function(e) NA)),
    ontology = sapply(go_id, function(id) tryCatch(GOTERM[[id]]@Ontology, error = function(e) NA)),
    is_deg   = gene_id %in% sig_genes
  )
head(go_long)
```

take GO terms, make dataframe for each treament enriched in and get plot for each treatment


```{r}
# ── 3. count genes per GO term, split by DEG status ─────────────────────────
go_counts <- go_long %>%
  filter(!is.na(term), !is.na(ontology)) %>%
  group_by(go_id, term, ontology) %>%
  summarise(
    n_total = n(),
    n_deg   = sum(is_deg),
    .groups = "drop"
  ) %>%
  filter(n_deg > 0) %>%                        # only keep terms with at least 1 DEG
  mutate(pct_deg = n_deg / n_total * 100)

# ── 4. plot: top N GO terms by number of DEGs, faceted by ontology ───────────
top_n_per_ontology <- 15   # adjust as needed

plot_data <- go_counts %>%
  group_by(ontology) %>%
  slice_max(n_deg, n = top_n_per_ontology) %>%
  ungroup() %>%
  mutate(term = stringr::str_wrap(term, width = 35),          # wrap long labels
         term = reorder(term, n_deg))

ggplot(plot_data, aes(x = term, y = n_deg, fill = ontology)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(n_deg, "/", n_total)),         # show deg/total
            hjust = -0.1, size = 3) +
  coord_flip() +
  facet_wrap(~ ontology, scales = "free_y",
             labeller = labeller(ontology = c(BP = "Biological Process",
                                              CC = "Cellular Component",
                                              MF = "Molecular Function"))) +
  scale_fill_manual(values = c(BP = "#4E79A7", CC = "#F28E2B", MF = "#59A14F")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +  # room for labels
  labs(
    title    = "GO Terms enriched in DEGs",
    subtitle = paste0("DEGs: padj < 0.05, |log2FC| > 1 (n = ", length(sig_genes), ")"),
    x        = NULL,
    y        = "Number of DEGs"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text  = element_text(face = "bold"),
        panel.grid.major.y = element_blank())
```





```{r}
res_aip_1dpi_ISOA
```


```{r}
tmp<-aip_annotate[match(rownames(mat_z), aip_annotate$gene_id),]
```






# 9.0 DESeq2 analysis on aggregated GOs

Structure: GOs --> SWISS Prot
1. Add annotations to count table 
2. Aggregate counts by GO string 
3. Run DESEQ2 on GO categories 
3. Visualize 










--> pick up here!
```{r helper for bicluster - GO categories}
# --- GO category assignment helpers (SwissProt first, TrEMBL fallback) ---

# Coalesce a GO column from SwissProt/TrEMBL
coalesce_go <- function(df, col_prefix) {
  sp <- paste0(col_prefix, "..SwissProt.")
  tr <- paste0(col_prefix, "..TrEMBL.")
  sp_vals <- if (sp %in% names(df)) df[[sp]] else NA_character_
  tr_vals <- if (tr %in% names(df)) df[[tr]] else NA_character_
  coalesce(sp_vals, tr_vals)
}

# Classify BP string into a broad category
categorize_bp <- function(s) {
  if (is.na(s) || s == "") return("Unannotated")
  s <- tolower(s)
  if (grepl("immune|defense|inflamm|complement|interleukin|cytokine|toll|antimicrobial", s)) return("Immune/defense")
  if (grepl("stress|oxidative|heat shock|apopto|caspase|programmed cell death", s))          return("Stress/apoptosis")
  if (grepl("transport|ion.*channel|membrane transport|vesicle", s))                         return("Transport")
  if (grepl("metabolic process|biosynth|catabol|lipid|fatty|carbohydrate", s))               return("Metabolism")
  if (grepl("sulfur|sulfo|sulfate|thiosulfate|methionine|DMSP|cysteine", s))                 return("Sulfur/S-cycling")
  if (grepl("cell cycle|division|mitot|DNA replication|repair", s))                          return("Cell cycle/DNA")
  if (grepl("signal|kinase|phosphoryl|receptor|G-protein", s))                               return("Signaling")
  if (grepl("translation|ribosom|rRNA|tRNA", s))                                             return("Translation")
  if (grepl("transcription|mRNA|splicing|chromatin|histone", s))                             return("Transcription/epigenetic")
  return("Other")
}

# Classify MF string
categorize_mf <- function(s) {
  if (is.na(s) || s == "") return("Unannotated")
  s <- tolower(s)
  if (grepl("kinase|phosphatase|phosphoryl", s))            return("Kinase/phosphatase")
  if (grepl("binding|receptor|ligand", s))                  return("Binding/receptor")
  if (grepl("oxidoreductase|dehydrogenase|oxidase|reductase|peroxidase", s)) return("Oxidoreductase")
  if (grepl("transferase", s))                              return("Transferase")
  if (grepl("hydrolase|protease|peptidase|nuclease", s))    return("Hydrolase/protease")
  if (grepl("transporter|channel|pump|carrier", s))         return("Transporter/channel")
  if (grepl("transcription factor|DNA.binding|zinc finger", s)) return("DNA-binding/TF")
  if (grepl("structural|cytoskel|actin|tubulin", s))        return("Structural")
  return("Other")
}

# Classify CC string
categorize_cc <- function(s) {
  if (is.na(s) || s == "") return("Unannotated")
  s <- tolower(s)
  if (grepl("membrane|integral|transmembrane", s))           return("Membrane")
  if (grepl("nucleus|nuclear|chromatin", s))                 return("Nucleus")
  if (grepl("cytoplasm|cytosol", s))                         return("Cytoplasm")
  if (grepl("mitochond", s))                                 return("Mitochondrion")
  if (grepl("extracellular|secreted", s))                    return("Extracellular")
  if (grepl("endoplasmic|golgi|lysosom|vacuol|vesicle", s))  return("Endomembrane")
  return("Other")
}

assign_bp_cat <- function(gene_ids) {
  bp_lookup <- aip_annotate %>%
    filter(gene_id %in% gene_ids) %>%
    mutate(bp_raw = coalesce_go(., "GO_BP"))
  cats <- sapply(bp_lookup$bp_raw, categorize_bp)
  setNames(cats, bp_lookup$gene_id)[gene_ids]
}

# Build all three GO category columns for a set of gene_ids
assign_go_cats <- function(gene_ids) {
  go_lookup <- aip_annotate %>%
    filter(gene_id %in% gene_ids) %>%
    mutate(bp_raw = coalesce_go(., "GO_BP"),
           mf_raw = coalesce_go(., "GO_MF"),
           cc_raw = coalesce_go(., "GO_CC"),
           BP = sapply(bp_raw, categorize_bp),
           MF = sapply(mf_raw, categorize_mf),
           CC = sapply(cc_raw, categorize_cc)) %>%
    dplyr::select(gene_id, BP, MF, CC)
  # Return as df with gene_ids in original order
  tibble(gene_id = gene_ids) %>% left_join(go_lookup, by = "gene_id") %>%
    mutate(across(c(BP, MF, CC), ~ replace_na(., "Unannotated")))
}

# --- GO annotation color palettes ---
# BP: warm reds/oranges/golds
# MF: cool purples/magentas
# CC: earthy greens/teals
# Unannotated = white, Other = light grey in all three

bp_colors <- c(
  "Immune/defense"           = "#B07A45",   
  "Stress/apoptosis"         = "#7A4019",   
  "Transport"                = "#E02020",   
  "Metabolism"               = "#E8701A",   
  "Sulfur/S-cycling"         = "#F5973A",   
  "Cell cycle/DNA"           = "#F5C242",   
  "Signaling"                = "#FFCBA4",   
  "Translation"              = "#FFB3C6",   
  "Transcription/epigenetic" = "#FF6EB4",   
  "Other"                    = "#D9D9D9",
  "Unannotated"              = "white")

mf_colors <- c(
  "Kinase/phosphatase"   = "#08306B",  
  "Binding/receptor"     = "#4292C6",  
  "Oxidoreductase"       = "#92C7E8",  
  "Transferase"          = "#74C476",  
  "Hydrolase/protease"   = "#D9A8E8",  
  "Transporter/channel"  = "#7424A3",   
  "DNA-binding/TF"       = "#3C0870",   
  "Structural"           = "#DF65B0",   
  "Other"                = "#D9D9D9",
  "Unannotated"          = "white")

cc_colors <- c(
  "Membrane"       = "#2E8B2E",   
  "Nucleus"        = "#00441B",   
  "Cytoplasm"      = "#BFDF60",  
  "Mitochondrion"  = "#B0E8A8",   
  "Extracellular"  = "#FFF280",   
  "Endomembrane"   = "#FDD835",  
  "Other"          = "#D9D9D9",
  "Unannotated"    = "white")

# Helper: build full annotation table for a set of gene_ids across multiple contrasts
build_deg_table <- function(gene_ids, contrasts_list) {
  fc_df <- map_dfr(names(contrasts_list), function(cn) {
    res <- as.data.frame(contrasts_list[[cn]]) %>% rownames_to_column("gene_id")
    res %>% filter(gene_id %in% gene_ids) %>%
      transmute(gene_id, contrast = cn, log2FoldChange, padj)
  })

  fc_wide <- fc_df %>%
    pivot_wider(names_from = contrast,
                values_from = c(log2FoldChange, padj),
                names_glue = "{contrast}_{.value}")

  annot_cols <- aip_annotate %>%
    dplyr::select(gene_id, protein_name,
           any_of(c("Hit_description..SwissProt.", "Hit_description..TrEMBL.",
                     "KEGG_ID")),
           starts_with("GO_BP"), starts_with("GO_MF"), starts_with("GO_CC"),
           starts_with("GO_IDs"))

  go_cats <- assign_go_cats(gene_ids)

  out <- annot_cols %>%
    filter(gene_id %in% gene_ids) %>%
    left_join(fc_wide, by = "gene_id") %>%
    left_join(go_cats, by = "gene_id") %>%
    dplyr::rename(BP_category = BP, MF_category = MF, CC_category = CC) %>%
    arrange(BP_category, gene_id)
  out
}
```

```{r help functions for plotting}
# Helper to build a 4A heatmap with BP + MF + CC row annotations and gene names
make_deg_heatmap <- function(sig_genes, vsd, timepoint_label) {
  mat <- assay(vsd[, vsd$timepoint == timepoint_label])[sig_genes, ]
  mat_z <- t(scale(t(mat)))
  mat_z <- mat_z[complete.cases(mat_z), ]
  cat(timepoint_label, "union DEGs:", length(sig_genes), "->", nrow(mat_z), "plotted\n")

  # Column annotation
  ann_col <- data.frame(
    Treatment = vsd[, vsd$timepoint == timepoint_label]$treatment,
    row.names = colnames(mat_z))

  # Row annotations: BP, MF, CC
  go_cats <- assign_go_cats(rownames(mat_z))
  ann_row <- data.frame(
    BP = go_cats$BP, MF = go_cats$MF, CC = go_cats$CC,
    row.names = rownames(mat_z))

  # Row labels: protein name, fallback to gene_id
  row_lab <- aip_annotate$protein_name[match(rownames(mat_z), aip_annotate$gene_id)]
  row_lab[is.na(row_lab) | row_lab == ""] <- rownames(mat_z)[is.na(row_lab) | row_lab == ""]
  row_lab <- make.unique(row_lab)

  p <- pheatmap(mat_z,
    annotation_col = ann_col,
    annotation_row = ann_row,
    annotation_colors = list(
      Treatment = palette,
      BP = bp_colors[intersect(names(bp_colors), unique(ann_row$BP))],
      MF = mf_colors[intersect(names(mf_colors), unique(ann_row$MF))],
      CC = cc_colors[intersect(names(cc_colors), unique(ann_row$CC))]),
    show_rownames = TRUE,
    labels_row = row_lab,
    fontsize_row = 5.5,
    cutree_rows = 3,
    cutree_cols = 4,
    main = paste0("Fig 4A: Union DEGs, Aiptasia ", timepoint_label,
                  " (", nrow(mat_z), " genes)"),
    silent = TRUE)

  grid::grid.newpage(); grid::grid.draw(p$gtable)

  list(pheatmap_obj = p, mat_z = mat_z, sig_genes = rownames(mat_z))
}
```

```{r plotting}
sig_1dpi_union <- unique(c(
  rownames(subset(as.data.frame(res_aip_1dpi_ISOA), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOB), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOC), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOABC), padj < 0.05))
))

sig<-results_1dpi %>% filter(padj<0.05) %>% pull(gene_id) %>% unique()

out_1dpi <- make_deg_heatmap(sig_1dpi_union, vsd_aip, "1dpi")
```

#####21 dpi
```{r plotting}
sig_21dpi_union <- unique(c(
  rownames(subset(as.data.frame(res_aip_21dpi_ISOA), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOB), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOC), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOABC), padj < 0.05))
))
out_21dpi <- make_deg_heatmap(sig_21dpi_union, vsd_aip, "1dpi")
```











# Archive


```{r}
ds$group <- factor(paste0(ds$timepoint, ds$treatment))
design(ds) <- ~ group
dds <- DESeq(ds)
resultsNames(dds)
colData(dds)
```
```{r}
unique(colData(dds)$group)
```

```{r get list of contrasts}
contrasts_1dpi<-data.frame(
  group1=c("1dpiABS","1dpiABS","1dpiABS","1dpiABS","1dpiABS","1dpimulti","1dpimulti","1dpimulti"),
  group2=c("1dpiLabrenzia", "1dpiRuegeria","1dpiVibrio", "1dpimulti","1dpiFASW","1dpiLabrenzia", "1dpiRuegeria","1dpiVibrio")
)

contrasts_21dpi<-data.frame(
  group1=c("21dpiABS","21dpiABS","21dpiABS","21dpiABS","21dpiABS","21dpimulti","21dpimulti","21dpimulti"),
  group2=c("21dpiLabrenzia", "21dpiRuegeria","21dpiVibrio", "21dpimulti","21dpiFASW","21dpiLabrenzia", "21dpiRuegeria","21dpiVibrio")
)

#contrast each treatment group bewteen day 1 and day 21
#contrasts_time<-data.frame(
#  group1=c("1dpiLabrenzia", "1dpiRuegeria","1dpiVibrio", "1dpimulti","1dpiFASW","1dpiABS"),
#  group2=c("21dpiLabrenzia", "21dpiRuegeria","21dpiVibrio", "21dpimulti","21dpiFASW","21dpiABS")
#)
#based on initial analyses, it doesn't make sense to compare day 1 and day 21 given that time is the dominant impact on gene expression

df_contrasts<-rbind(contrasts_1dpi,contrasts_21dpi)
df_contrasts
```
```{r get all pairwise contrast results}
results_list <- lapply(1:nrow(df_contrasts), function(i) {
  g1 <- df_contrasts$group1[i]
  g2 <- df_contrasts$group2[i]
  res <- results(dds, contrast = c("group", g1, g2))
  df  <- as.data.frame(res)
  df$gene_id  <- rownames(df)
  df$contrast <- paste(g1, "vs", g2,sep="_")
  return(df)
}) |>
  bind_rows()
results_list
```

```{r}
length(unique(results_list$contrast))
results_list
```

```{r get summary table}
deg_summary<-results_list %>% group_by(contrast) %>% 
                 summarize(up = sum(log2FoldChange > 1 & padj<0.05, na.rm = TRUE), 
                           down=sum(log2FoldChange < 1& padj<0.05, na.rm = TRUE),
                           total_deg=sum(padj<0.05 & abs(log2FoldChange)>1, na.rm=T))

deg_summary <- deg_summary |>
  mutate(
    # Extract timepoint from each side of the contrast
    timepoint1 = str_extract(str_extract(contrast, "^[^_]+"), "\\d+dpi"),
    timepoint2 = str_extract(str_extract(contrast, "(?<=_vs_).+$"), "\\d+dpi"),
    
    # Extract treatment from each side
    treatment1 = str_remove(str_extract(contrast, "^[^_]+"), "\\d+dpi"),
    treatment2 = str_remove(str_extract(contrast, "(?<=_vs_).+$"), "\\d+dpi"),
    
    # Classify contrast type
    contrast_type = case_when(
      timepoint1 == timepoint2 ~ "within_timepoint",
      treatment1 == treatment2 ~ "across_timepoint",
      TRUE ~ "mixed"
    ),
    
    # Assign a shared timepoint or treatment label for grouping
    timepoint_group = case_when(
      timepoint1 == timepoint2 ~ timepoint1,
      TRUE ~ paste(timepoint1, "vs", timepoint2)
    ),
    treatment_group = case_when(
      treatment1 == treatment2 ~ treatment1,
      TRUE ~ paste(treatment1, "vs", treatment2)
    )
  )
deg_summary
```

### plot nr of up and down regulated genes
```{r}
deg_long <- deg_summary |>
  dplyr::select(contrast, contrast_type, treatment_group, timepoint_group, up, down) |>
  mutate(down = -down) |>  # negative for diverging bar
  pivot_longer(cols = c(up, down), names_to = "direction", values_to = "count")

ggplot(deg_long, aes(x=treatment_group, y=count, fill=direction)) + 
      geom_bar(stat="identity") + 
      facet_wrap(~timepoint_group,scales='free_x') + 
      theme_bw() + 
      theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
```

### get deg table and annotation list
```{r filter result lists and add annotations}
deg_aip<-results_list %>% 
                filter(padj< 0.05, abs(log2FoldChange)>1) %>% 
                dplyr::select(contrast,gene_id, padj,log2FoldChange) %>% 
                left_join(aip_annotate, by="gene_id")
head(deg_aip); dim(deg_aip)
```

```{r export}
write.csv(deg_aip, file.path(path.out, "DEG_all_aip.csv"), row.names = FALSE)
saveRDS(dds, file.path(path.out,"dds_aip.rds"))
```

## 8.6 Visualization 

### bi-cluster heatmap (VST, Z-scored)
```{r get vst data}
sig.genes<-unique(deg_aip$gene_id); length(sig.genes)
vst.dds <- vst(dds, blind = TRUE)
vst.data<-assay(vst.dds)
vst.de<-vst.data[rownames(vst.data) %in% sig.genes,]
dim(vst.de)
```

```{r get metadata}
col.ann <- as.data.frame(colData(dds)) |>
  dplyr::select(timepoint, treatment)  # adjust column names to match yours

rownames(col.ann) <- colnames(vst.de)
```
```{r}
# Colour palette
ann_colors <- list(
  timepoint = c("1dpi"  = "#5DCAA5",
                "21dpi" = "#085041"),
  treatment = c("ABS"       = "black",
                "FASW"      = "grey60",
                "Labrenzia" = "#A6CEE3",
                "Ruegeria"  = "#1F78B4",
                "Vibrio"    = "#B2DF8A",
                "multi"     = "violet")
)

pheatmap(
  vst.de,
  annotation_col  = col.ann,
  annotation_colors = ann_colors,
  scale           = "row",        # z-score per gene
  show_rownames   = FALSE,        # too many genes to label
  show_colnames   = FALSE,
  cluster_rows    = TRUE,
  cluster_cols    = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  color           = colorRampPalette(c("#2a78d6", "white", "#e34948"))(100),
  border_color    = NA,
  fontsize        = 10,
  main            = "DE genes — VST scaled expression",
  cutree_rows     = 3,
  cutree_cols     = 4)
```





```{r}
  p <- pheatmap(mat_z,
    annotation_col = ann_col,
    annotation_row = ann_row,
    annotation_colors = list(
      Treatment = palette,
      BP = bp_colors[intersect(names(bp_colors), unique(ann_row$BP))],
      MF = mf_colors[intersect(names(mf_colors), unique(ann_row$MF))],
      CC = cc_colors[intersect(names(cc_colors), unique(ann_row$CC))]),
    show_rownames = TRUE,
    labels_row = row_lab,
    fontsize_row = 5.5,
    cutree_rows = 3,
    cutree_cols = 4,
    main = paste0("Fig 4A: Union DEGs, Aiptasia ", timepoint_label,
                  " (", nrow(mat_z), " genes)"),
    silent = TRUE)


```







Union of all DE genes from treatment vs ABS contrasts, shown as Z-scored VST
heatmap with GO category annotations (BP, MF, CC).

```{r fig4a-helper}
# --- GO category assignment helpers (SwissProt first, TrEMBL fallback) ---

# Coalesce a GO column from SwissProt/TrEMBL
coalesce_go <- function(df, col_prefix) {
  sp <- paste0(col_prefix, "..SwissProt.")
  tr <- paste0(col_prefix, "..TrEMBL.")
  sp_vals <- if (sp %in% names(df)) df[[sp]] else NA_character_
  tr_vals <- if (tr %in% names(df)) df[[tr]] else NA_character_
  coalesce(sp_vals, tr_vals)
}

# Classify BP string into a broad category
categorize_bp <- function(s) {
  if (is.na(s) || s == "") return("Unannotated")
  s <- tolower(s)
  if (grepl("immune|defense|inflamm|complement|interleukin|cytokine|toll|antimicrobial", s)) return("Immune/defense")
  if (grepl("stress|oxidative|heat shock|apopto|caspase|programmed cell death", s))          return("Stress/apoptosis")
  if (grepl("transport|ion.*channel|membrane transport|vesicle", s))                         return("Transport")
  if (grepl("metabolic process|biosynth|catabol|lipid|fatty|carbohydrate", s))               return("Metabolism")
  if (grepl("sulfur|sulfo|sulfate|thiosulfate|methionine|DMSP|cysteine", s))                 return("Sulfur/S-cycling")
  if (grepl("cell cycle|division|mitot|DNA replication|repair", s))                          return("Cell cycle/DNA")
  if (grepl("signal|kinase|phosphoryl|receptor|G-protein", s))                               return("Signaling")
  if (grepl("translation|ribosom|rRNA|tRNA", s))                                             return("Translation")
  if (grepl("transcription|mRNA|splicing|chromatin|histone", s))                             return("Transcription/epigenetic")
  return("Other")
}

# Classify MF string
categorize_mf <- function(s) {
  if (is.na(s) || s == "") return("Unannotated")
  s <- tolower(s)
  if (grepl("kinase|phosphatase|phosphoryl", s))            return("Kinase/phosphatase")
  if (grepl("binding|receptor|ligand", s))                  return("Binding/receptor")
  if (grepl("oxidoreductase|dehydrogenase|oxidase|reductase|peroxidase", s)) return("Oxidoreductase")
  if (grepl("transferase", s))                              return("Transferase")
  if (grepl("hydrolase|protease|peptidase|nuclease", s))    return("Hydrolase/protease")
  if (grepl("transporter|channel|pump|carrier", s))         return("Transporter/channel")
  if (grepl("transcription factor|DNA.binding|zinc finger", s)) return("DNA-binding/TF")
  if (grepl("structural|cytoskel|actin|tubulin", s))        return("Structural")
  return("Other")
}

# Classify CC string
categorize_cc <- function(s) {
  if (is.na(s) || s == "") return("Unannotated")
  s <- tolower(s)
  if (grepl("membrane|integral|transmembrane", s))           return("Membrane")
  if (grepl("nucleus|nuclear|chromatin", s))                 return("Nucleus")
  if (grepl("cytoplasm|cytosol", s))                         return("Cytoplasm")
  if (grepl("mitochond", s))                                 return("Mitochondrion")
  if (grepl("extracellular|secreted", s))                    return("Extracellular")
  if (grepl("endoplasmic|golgi|lysosom|vacuol|vesicle", s))  return("Endomembrane")
  return("Other")
}

assign_bp_cat <- function(gene_ids) {
  bp_lookup <- aip_annotate %>%
    filter(gene_id %in% gene_ids) %>%
    mutate(bp_raw = coalesce_go(., "GO_BP"))
  cats <- sapply(bp_lookup$bp_raw, categorize_bp)
  setNames(cats, bp_lookup$gene_id)[gene_ids]
}

# Build all three GO category columns for a set of gene_ids
assign_go_cats <- function(gene_ids) {
  go_lookup <- aip_annotate %>%
    filter(gene_id %in% gene_ids) %>%
    mutate(bp_raw = coalesce_go(., "GO_BP"),
           mf_raw = coalesce_go(., "GO_MF"),
           cc_raw = coalesce_go(., "GO_CC"),
           BP = sapply(bp_raw, categorize_bp),
           MF = sapply(mf_raw, categorize_mf),
           CC = sapply(cc_raw, categorize_cc)) %>%
    select(gene_id, BP, MF, CC)
  # Return as df with gene_ids in original order
  tibble(gene_id = gene_ids) %>% left_join(go_lookup, by = "gene_id") %>%
    mutate(across(c(BP, MF, CC), ~ replace_na(., "Unannotated")))
}

# --- GO annotation color palettes ---
# BP: warm reds/oranges/golds
# MF: cool purples/magentas
# CC: earthy greens/teals
# Unannotated = white, Other = light grey in all three

bp_colors <- c(
  "Immune/defense"           = "#B07A45",   
  "Stress/apoptosis"         = "#7A4019",   
  "Transport"                = "#E02020",   
  "Metabolism"               = "#E8701A",   
  "Sulfur/S-cycling"         = "#F5973A",   
  "Cell cycle/DNA"           = "#F5C242",   
  "Signaling"                = "#FFCBA4",   
  "Translation"              = "#FFB3C6",   
  "Transcription/epigenetic" = "#FF6EB4",   
  "Other"                    = "#D9D9D9",
  "Unannotated"              = "white")

mf_colors <- c(
  "Kinase/phosphatase"   = "#08306B",  
  "Binding/receptor"     = "#4292C6",  
  "Oxidoreductase"       = "#92C7E8",  
  "Transferase"          = "#74C476",  
  "Hydrolase/protease"   = "#D9A8E8",  
  "Transporter/channel"  = "#7424A3",   
  "DNA-binding/TF"       = "#3C0870",   
  "Structural"           = "#DF65B0",   
  "Other"                = "#D9D9D9",
  "Unannotated"          = "white")


cc_colors <- c(
  "Membrane"       = "#2E8B2E",   
  "Nucleus"        = "#00441B",   
  "Cytoplasm"      = "#BFDF60",  
  "Mitochondrion"  = "#B0E8A8",   
  "Extracellular"  = "#FFF280",   
  "Endomembrane"   = "#FDD835",  
  "Other"          = "#D9D9D9",
  "Unannotated"    = "white")

# Helper: build full annotation table for a set of gene_ids across multiple contrasts
build_deg_table <- function(gene_ids, contrasts_list) {
  fc_df <- map_dfr(names(contrasts_list), function(cn) {
    res <- as.data.frame(contrasts_list[[cn]]) %>% rownames_to_column("gene_id")
    res %>% filter(gene_id %in% gene_ids) %>%
      transmute(gene_id, contrast = cn, log2FoldChange, padj)
  })

  fc_wide <- fc_df %>%
    pivot_wider(names_from = contrast,
                values_from = c(log2FoldChange, padj),
                names_glue = "{contrast}_{.value}")

  annot_cols <- aip_annotate %>%
    select(gene_id, protein_name,
           any_of(c("Hit_description..SwissProt.", "Hit_description..TrEMBL.",
                     "KEGG_ID")),
           starts_with("GO_BP"), starts_with("GO_MF"), starts_with("GO_CC"),
           starts_with("GO_IDs"))

  go_cats <- assign_go_cats(gene_ids)

  out <- annot_cols %>%
    filter(gene_id %in% gene_ids) %>%
    left_join(fc_wide, by = "gene_id") %>%
    left_join(go_cats, by = "gene_id") %>%
    dplyr::rename(BP_category = BP, MF_category = MF, CC_category = CC) %>%
    arrange(BP_category, gene_id)
  out
}
```






## 8.6 Volcano Plot Helper + Downregulated Gene Lists

### Volcano Plot Helper Function

```{r volcano-setup}
plot_volcano <- function(res, annotation, title) {
  df <- as.data.frame(res) %>%
    rownames_to_column("gene_id") %>%
    left_join(annotation, by = "gene_id") %>%
    mutate(sig = case_when(
      padj < 0.05 & log2FoldChange > 1  ~ "Up",
      padj < 0.05 & log2FoldChange < -1 ~ "Down",
      TRUE ~ "NS"))

  top_genes <- df %>% filter(sig != "NS") %>%
    slice_min(padj, n = 10, with_ties = FALSE)

  ggplot(df, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
    geom_point(alpha = 0.4, size = 1) +
    scale_color_manual(values = c("Up" = "red", "Down" = "blue", "NS" = "gray70")) +
    geom_text_repel(data = top_genes, aes(label = protein_name),
                    size = 3, max.overlaps = 15) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray40") +
    theme_bw() +
    labs(title = title, x = "log2 Fold Change", y = "-log10(adjusted p-value)")
}
```

### Downregulated Gene Lists

Extract downregulated DEGs from each host contrast (padj<0.05, log2FC<-1),
annotate with descriptions and GO/KEGG to see what functional categories
are being suppressed.

```{r volcano-downregulated}
all_host_contrasts_vol <- list(
  "1dpi_FASW_vs_ABS"       = res_aip_1dpi_control,
  "1dpi_Labrenzia_vs_ABS"  = res_aip_1dpi_ISOA,
  "1dpi_Ruegeria_vs_ABS"   = res_aip_1dpi_ISOB,
  "1dpi_Vibrio_vs_ABS"     = res_aip_1dpi_ISOC,
  "1dpi_multi_vs_ABS"      = res_aip_1dpi_ISOABC,
  "21dpi_FASW_vs_ABS"      = res_aip_21dpi_control,
  "21dpi_Labrenzia_vs_ABS" = res_aip_21dpi_ISOA,
  "21dpi_Ruegeria_vs_ABS"  = res_aip_21dpi_ISOB,
  "21dpi_Vibrio_vs_ABS"    = res_aip_21dpi_ISOC,
  "21dpi_multi_vs_ABS"     = res_aip_21dpi_ISOABC
)

down_genes_all <- map_dfr(names(all_host_contrasts_vol), function(cn) {
  res <- all_host_contrasts_vol[[cn]]
  as.data.frame(res) %>%
    rownames_to_column("gene_id") %>%
    filter(!is.na(padj), padj < 0.05, log2FoldChange < -1) %>%
    arrange(log2FoldChange) %>%
    left_join(aip_annotate %>%
                dplyr::select(gene_id, protein_name,
                       any_of(c("Hit_description..SwissProt.",
                                "Hit_description..TrEMBL.",
                                "KEGG_ID")),
                       starts_with("GO_IDs")),
              by = "gene_id") %>%
    mutate(contrast = cn) %>%
    dplyr::select(contrast, gene_id, protein_name, log2FoldChange, padj, everything())
})

down_genes_all %>% dplyr::count(contrast, name = "n_downregulated") %>% print()

write_csv(down_genes_all, "../results/host_downregulated_genes_all_contrasts.csv")
cat("Full table saved to ../results/host_downregulated_genes_all_contrasts.csv\n")
```


## 8.8 Fig 4: Host DE Response to Bacterial Inoculation

### Fig 4A — Bicluster Heatmap (VST, Z-scored)

Union of all DE genes from treatment vs ABS contrasts, shown as Z-scored VST
heatmap with GO category annotations (BP, MF, CC).

```{r fig4a-helper}
# --- GO category assignment helpers (SwissProt first, TrEMBL fallback) ---

# Coalesce a GO column from SwissProt/TrEMBL
coalesce_go <- function(df, col_prefix) {
  sp <- paste0(col_prefix, "..SwissProt.")
  tr <- paste0(col_prefix, "..TrEMBL.")
  sp_vals <- if (sp %in% names(df)) df[[sp]] else NA_character_
  tr_vals <- if (tr %in% names(df)) df[[tr]] else NA_character_
  coalesce(sp_vals, tr_vals)
}

# Classify BP string into a broad category
categorize_bp <- function(s) {
  if (is.na(s) || s == "") return("Unannotated")
  s <- tolower(s)
  if (grepl("immune|defense|inflamm|complement|interleukin|cytokine|toll|antimicrobial", s)) return("Immune/defense")
  if (grepl("stress|oxidative|heat shock|apopto|caspase|programmed cell death", s))          return("Stress/apoptosis")
  if (grepl("transport|ion.*channel|membrane transport|vesicle", s))                         return("Transport")
  if (grepl("metabolic process|biosynth|catabol|lipid|fatty|carbohydrate", s))               return("Metabolism")
  if (grepl("sulfur|sulfo|sulfate|thiosulfate|methionine|DMSP|cysteine", s))                 return("Sulfur/S-cycling")
  if (grepl("cell cycle|division|mitot|DNA replication|repair", s))                          return("Cell cycle/DNA")
  if (grepl("signal|kinase|phosphoryl|receptor|G-protein", s))                               return("Signaling")
  if (grepl("translation|ribosom|rRNA|tRNA", s))                                             return("Translation")
  if (grepl("transcription|mRNA|splicing|chromatin|histone", s))                             return("Transcription/epigenetic")
  return("Other")
}

# Classify MF string
categorize_mf <- function(s) {
  if (is.na(s) || s == "") return("Unannotated")
  s <- tolower(s)
  if (grepl("kinase|phosphatase|phosphoryl", s))            return("Kinase/phosphatase")
  if (grepl("binding|receptor|ligand", s))                  return("Binding/receptor")
  if (grepl("oxidoreductase|dehydrogenase|oxidase|reductase|peroxidase", s)) return("Oxidoreductase")
  if (grepl("transferase", s))                              return("Transferase")
  if (grepl("hydrolase|protease|peptidase|nuclease", s))    return("Hydrolase/protease")
  if (grepl("transporter|channel|pump|carrier", s))         return("Transporter/channel")
  if (grepl("transcription factor|DNA.binding|zinc finger", s)) return("DNA-binding/TF")
  if (grepl("structural|cytoskel|actin|tubulin", s))        return("Structural")
  return("Other")
}

# Classify CC string
categorize_cc <- function(s) {
  if (is.na(s) || s == "") return("Unannotated")
  s <- tolower(s)
  if (grepl("membrane|integral|transmembrane", s))           return("Membrane")
  if (grepl("nucleus|nuclear|chromatin", s))                 return("Nucleus")
  if (grepl("cytoplasm|cytosol", s))                         return("Cytoplasm")
  if (grepl("mitochond", s))                                 return("Mitochondrion")
  if (grepl("extracellular|secreted", s))                    return("Extracellular")
  if (grepl("endoplasmic|golgi|lysosom|vacuol|vesicle", s))  return("Endomembrane")
  return("Other")
}

assign_bp_cat <- function(gene_ids) {
  bp_lookup <- aip_annotate %>%
    filter(gene_id %in% gene_ids) %>%
    mutate(bp_raw = coalesce_go(., "GO_BP"))
  cats <- sapply(bp_lookup$bp_raw, categorize_bp)
  setNames(cats, bp_lookup$gene_id)[gene_ids]
}

# Build all three GO category columns for a set of gene_ids
assign_go_cats <- function(gene_ids) {
  go_lookup <- aip_annotate %>%
    filter(gene_id %in% gene_ids) %>%
    mutate(bp_raw = coalesce_go(., "GO_BP"),
           mf_raw = coalesce_go(., "GO_MF"),
           cc_raw = coalesce_go(., "GO_CC"),
           BP = sapply(bp_raw, categorize_bp),
           MF = sapply(mf_raw, categorize_mf),
           CC = sapply(cc_raw, categorize_cc)) %>%
    select(gene_id, BP, MF, CC)
  # Return as df with gene_ids in original order
  tibble(gene_id = gene_ids) %>% left_join(go_lookup, by = "gene_id") %>%
    mutate(across(c(BP, MF, CC), ~ replace_na(., "Unannotated")))
}

# --- GO annotation color palettes ---
# BP: warm reds/oranges/golds
# MF: cool purples/magentas
# CC: earthy greens/teals
# Unannotated = white, Other = light grey in all three

bp_colors <- c(
  "Immune/defense"           = "#B07A45",   
  "Stress/apoptosis"         = "#7A4019",   
  "Transport"                = "#E02020",   
  "Metabolism"               = "#E8701A",   
  "Sulfur/S-cycling"         = "#F5973A",   
  "Cell cycle/DNA"           = "#F5C242",   
  "Signaling"                = "#FFCBA4",   
  "Translation"              = "#FFB3C6",   
  "Transcription/epigenetic" = "#FF6EB4",   
  "Other"                    = "#D9D9D9",
  "Unannotated"              = "white")

mf_colors <- c(
  "Kinase/phosphatase"   = "#08306B",  
  "Binding/receptor"     = "#4292C6",  
  "Oxidoreductase"       = "#92C7E8",  
  "Transferase"          = "#74C476",  
  "Hydrolase/protease"   = "#D9A8E8",  
  "Transporter/channel"  = "#7424A3",   
  "DNA-binding/TF"       = "#3C0870",   
  "Structural"           = "#DF65B0",   
  "Other"                = "#D9D9D9",
  "Unannotated"          = "white")


cc_colors <- c(
  "Membrane"       = "#2E8B2E",   
  "Nucleus"        = "#00441B",   
  "Cytoplasm"      = "#BFDF60",  
  "Mitochondrion"  = "#B0E8A8",   
  "Extracellular"  = "#FFF280",   
  "Endomembrane"   = "#FDD835",  
  "Other"          = "#D9D9D9",
  "Unannotated"    = "white")

# Helper: build full annotation table for a set of gene_ids across multiple contrasts
build_deg_table <- function(gene_ids, contrasts_list) {
  fc_df <- map_dfr(names(contrasts_list), function(cn) {
    res <- as.data.frame(contrasts_list[[cn]]) %>% rownames_to_column("gene_id")
    res %>% filter(gene_id %in% gene_ids) %>%
      transmute(gene_id, contrast = cn, log2FoldChange, padj)
  })

  fc_wide <- fc_df %>%
    pivot_wider(names_from = contrast,
                values_from = c(log2FoldChange, padj),
                names_glue = "{contrast}_{.value}")

  annot_cols <- aip_annotate %>%
    select(gene_id, protein_name,
           any_of(c("Hit_description..SwissProt.", "Hit_description..TrEMBL.",
                     "KEGG_ID")),
           starts_with("GO_BP"), starts_with("GO_MF"), starts_with("GO_CC"),
           starts_with("GO_IDs"))

  go_cats <- assign_go_cats(gene_ids)

  out <- annot_cols %>%
    filter(gene_id %in% gene_ids) %>%
    left_join(fc_wide, by = "gene_id") %>%
    left_join(go_cats, by = "gene_id") %>%
    dplyr::rename(BP_category = BP, MF_category = MF, CC_category = CC) %>%
    arrange(BP_category, gene_id)
  out
}
```

```{r fig4a-bicluster-heatmap, fig.width=14, fig.height=16}
# Helper to build a 4A heatmap with BP + MF + CC row annotations and gene names
make_fig4a <- function(sig_genes, vsd, timepoint_label) {
  mat <- assay(vsd[, vsd$timepoint == timepoint_label])[sig_genes, ]
  mat_z <- t(scale(t(mat)))
  mat_z <- mat_z[complete.cases(mat_z), ]
  cat(timepoint_label, "union DEGs:", length(sig_genes), "->", nrow(mat_z), "plotted\n")

  # Column annotation
  ann_col <- data.frame(
    Treatment = vsd[, vsd$timepoint == timepoint_label]$treatment,
    row.names = colnames(mat_z))

  # Row annotations: BP, MF, CC
  go_cats <- assign_go_cats(rownames(mat_z))
  ann_row <- data.frame(
    BP = go_cats$BP, MF = go_cats$MF, CC = go_cats$CC,
    row.names = rownames(mat_z))

  # Row labels: protein name, fallback to gene_id
  row_lab <- aip_annotate$protein_name[match(rownames(mat_z), aip_annotate$gene_id)]
  row_lab[is.na(row_lab) | row_lab == ""] <- rownames(mat_z)[is.na(row_lab) | row_lab == ""]
  row_lab <- make.unique(row_lab)

  p <- pheatmap(mat_z,
    annotation_col = ann_col,
    annotation_row = ann_row,
    annotation_colors = list(
      Treatment = palette,
      BP = bp_colors[intersect(names(bp_colors), unique(ann_row$BP))],
      MF = mf_colors[intersect(names(mf_colors), unique(ann_row$MF))],
      CC = cc_colors[intersect(names(cc_colors), unique(ann_row$CC))]),
    show_rownames = TRUE,
    labels_row = row_lab,
    fontsize_row = 5.5,
    cutree_rows = 3,
    cutree_cols = 4,
    main = paste0("Fig 4A: Union DEGs, Aiptasia ", timepoint_label,
                  " (", nrow(mat_z), " genes)"),
    silent = TRUE)

  grid::grid.newpage(); grid::grid.draw(p$gtable)

  list(pheatmap_obj = p, mat_z = mat_z, sig_genes = rownames(mat_z))
}

# --- 1dpi ---
sig_1dpi_union <- unique(c(
  rownames(subset(as.data.frame(res_aip_1dpi_ISOA), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOB), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOC), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOABC), padj < 0.05))
))
out_1dpi <- make_fig4a(sig_1dpi_union, vsd_aip, "1dpi")
fig4a_1dpi <- out_1dpi$pheatmap_obj

# --- 21dpi ---
sig_21dpi_union <- unique(c(
  rownames(subset(as.data.frame(res_aip_21dpi_ISOA), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOB), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOC), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOABC), padj < 0.05))
))
out_21dpi <- make_fig4a(sig_21dpi_union, vsd_aip, "21dpi")
fig4a_21dpi <- out_21dpi$pheatmap_obj
```

### Fig 4A — Full Annotated DEG Tables

Searchable table of **every** DEG from the Fig 4A union sets, with gene name,
GO BP/MF, KEGG, SwissProt description, and log2FC + padj per contrast.

```{r fig4a-deg-tables}
fig4a_contrasts_1dpi <- list(
  Labrenzia_vs_ABS = res_aip_1dpi_ISOA,
  Ruegeria_vs_ABS  = res_aip_1dpi_ISOB,
  Vibrio_vs_ABS    = res_aip_1dpi_ISOC,
  multi_vs_ABS     = res_aip_1dpi_ISOABC
)
fig4a_contrasts_21dpi <- list(
  Labrenzia_vs_ABS = res_aip_21dpi_ISOA,
  Ruegeria_vs_ABS  = res_aip_21dpi_ISOB,
  Vibrio_vs_ABS    = res_aip_21dpi_ISOC,
  multi_vs_ABS     = res_aip_21dpi_ISOABC
)

fig4a_table_1dpi  <- build_deg_table(sig_1dpi_union, fig4a_contrasts_1dpi)
fig4a_table_21dpi <- build_deg_table(sig_21dpi_union, fig4a_contrasts_21dpi)

cat("Fig 4A 1dpi:", nrow(fig4a_table_1dpi), "genes\n")
cat("Fig 4A 21dpi:", nrow(fig4a_table_21dpi), "genes\n")

write_csv(fig4a_table_1dpi, "../results/Fig4A_DEG_table_1dpi.csv")
write_csv(fig4a_table_21dpi, "../results/Fig4A_DEG_table_21dpi.csv")

# Print a summary: gene counts per BP category
cat("\n--- 1dpi BP category breakdown ---\n")
fig4a_table_1dpi %>% dplyr::count(BP_category, sort = TRUE) %>% print()
cat("\n--- 21dpi BP category breakdown ---\n")
fig4a_table_21dpi %>% dplyr::count(BP_category, sort = TRUE) %>% print()

cat("Full tables saved to ../results/Fig4A_DEG_table_*.csv\n")
```

## 8.9 Fig 5: Single Isolate vs Community — Host Gene Response

Uses single vs community contrasts from Section 8.4. Includes Vibrio panels
to test whether community-enriched genes are truly community-specific
or driven by Vibrio presence. This is host data (how the host responds
differently to single inocula vs the 3-member community).

### Fig 5A-F — Volcano Plots (log2FC vs -log10 padj)

```{r fig5-volcanoes, fig.width=14, fig.height=16}
# Reuses plot_volcano() from Section 8.6
# positive log2FC = HIGHER in single inoculation
# negative log2FC = HIGHER in community (multi)
# Red/Up = enriched in single | Blue/Down = enriched in community

p_fig5a <- plot_volcano(res_aip_1dpi_ISOA_vs_ISOABC, aip_annotate,
            "5A: 1dpi Labrenzia vs Community")
p_fig5b <- plot_volcano(res_aip_21dpi_ISOA_vs_ISOABC, aip_annotate,
            "5B: 21dpi Labrenzia vs Community")
p_fig5c <- plot_volcano(res_aip_1dpi_ISOB_vs_ISOABC, aip_annotate,
            "5C: 1dpi Ruegeria vs Community")
p_fig5d <- plot_volcano(res_aip_21dpi_ISOB_vs_ISOABC, aip_annotate,
            "5D: 21dpi Ruegeria vs Community")
p_fig5e_vib <- plot_volcano(res_aip_1dpi_ISOC_vs_ISOABC, aip_annotate,
            "5E: 1dpi Vibrio vs Community")
p_fig5f_vib <- plot_volcano(res_aip_21dpi_ISOC_vs_ISOABC, aip_annotate,
            "5F: 21dpi Vibrio vs Community")

(p_fig5a + p_fig5b) / (p_fig5c + p_fig5d) / (p_fig5e_vib + p_fig5f_vib)
```

### Fig 5 — Gene Function Key

```{r fig5-gene-key}
# Collect the top-10-by-padj genes that get labeled on each volcano
fig5_key <- map_dfr(list(
  "5A: 1dpi Lab vs Community"  = res_aip_1dpi_ISOA_vs_ISOABC,
  "5B: 21dpi Lab vs Community" = res_aip_21dpi_ISOA_vs_ISOABC,
  "5C: 1dpi Rue vs Community"  = res_aip_1dpi_ISOB_vs_ISOABC,
  "5D: 21dpi Rue vs Community" = res_aip_21dpi_ISOB_vs_ISOABC,
  "5E: 1dpi Vib vs Community"  = res_aip_1dpi_ISOC_vs_ISOABC,
  "5F: 21dpi Vib vs Community" = res_aip_21dpi_ISOC_vs_ISOABC
), function(res) {
  as.data.frame(res) %>%
    rownames_to_column("gene_id") %>%
    filter(!is.na(padj)) %>%
    mutate(sig = case_when(
      padj < 0.05 & log2FoldChange > 1  ~ "Up (single)",
      padj < 0.05 & log2FoldChange < -1 ~ "Up (community)",
      TRUE ~ "NS")) %>%
    filter(sig != "NS") %>%
    slice_min(padj, n = 10, with_ties = FALSE)
}, .id = "panel") %>%
  left_join(aip_annotate %>%
    select(gene_id, protein_name, Hit_description..SwissProt.,
           any_of(c("GO_BP..SwissProt.", "GO_BP..TrEMBL."))),
    by = "gene_id") %>%
  mutate(
    bp_annot = coalesce(
      if ("GO_BP..SwissProt." %in% names(.)) .[["GO_BP..SwissProt."]] else NA_character_,
      if ("GO_BP..TrEMBL." %in% names(.)) .[["GO_BP..TrEMBL."]] else NA_character_),
    function_summary = case_when(
      !is.na(Hit_description..SwissProt.) & Hit_description..SwissProt. != "" ~
        str_trunc(Hit_description..SwissProt., 60),
      !is.na(bp_annot) & bp_annot != "" ~ str_trunc(bp_annot, 60),
      TRUE ~ "Unknown function")) %>%
  select(panel, gene_id, protein_name, direction = sig,
         log2FoldChange, padj, function_summary) %>%
  distinct(gene_id, panel, .keep_all = TRUE) %>%
  arrange(panel, direction, padj)

write_csv(fig5_key, "../results/Fig5_volcano_gene_key.csv")
cat("Fig 5 gene key:", nrow(fig5_key), "genes saved to ../results/Fig5_volcano_gene_key.csv\n")
```

### Fig 5 — Full Annotated DEG Table

```{r fig5-deg-table}
fig5_contrasts <- list(
  "1dpi_Labrenzia_vs_Community"  = res_aip_1dpi_ISOA_vs_ISOABC,
  "21dpi_Labrenzia_vs_Community" = res_aip_21dpi_ISOA_vs_ISOABC,
  "1dpi_Ruegeria_vs_Community"   = res_aip_1dpi_ISOB_vs_ISOABC,
  "21dpi_Ruegeria_vs_Community"  = res_aip_21dpi_ISOB_vs_ISOABC,
  "1dpi_Vibrio_vs_Community"     = res_aip_1dpi_ISOC_vs_ISOABC,
  "21dpi_Vibrio_vs_Community"    = res_aip_21dpi_ISOC_vs_ISOABC
)

fig5_all_sig <- unique(unlist(lapply(fig5_contrasts, function(res) {
  rownames(subset(as.data.frame(res), padj < 0.05))
})))
cat("Fig 5 total unique DEGs:", length(fig5_all_sig), "\n")

if (length(fig5_all_sig) > 0) {
  fig5_table <- build_deg_table(fig5_all_sig, fig5_contrasts)
  write_csv(fig5_table, "../results/Fig5_DEG_table_single_vs_community.csv")

  cat("\n--- Fig 5 BP category breakdown ---\n")
  fig5_table %>% dplyr::count(BP_category, sort = TRUE) %>% print()

  cat("Full table saved to ../results/Fig5_DEG_table_single_vs_community.csv\n")
} else {
  cat("No significant DEGs in any Fig 5 contrast\n")
}
```

## 8.10 Pathway Composition — DEG Counts by GO Category

Stacked bar charts showing the breakdown of up- and downregulated DEGs (padj < 0.05) into GO Biological Process, Molecular Function, and Cellular Component categories. Positive bars = upregulated vs ABS; negative bars = downregulated vs ABS.

```{r pathway-composition, fig.width=14, fig.height=14}
# Collect all DEGs from treatment vs ABS contrasts with GO categories
deg_lfc_pre <- map_dfr(names(all_host_contrasts_vol), function(cn) {
  as.data.frame(all_host_contrasts_vol[[cn]]) %>%
    rownames_to_column("gene_id") %>%
    filter(!is.na(padj), padj < 0.05) %>%
    mutate(contrast  = cn,
           direction = if_else(log2FoldChange > 0, "Up", "Down", missing = "Down"),
           timepoint = str_extract(cn, "^\\d+dpi"),
           treatment = str_extract(cn, "(?<=dpi_)[^_]+(?=_vs)"))
})

go_cats_all <- assign_go_cats(unique(deg_lfc_pre$gene_id))
deg_lfc_cats <- deg_lfc_pre %>%
  left_join(go_cats_all, by = "gene_id") %>%
  mutate(BP = as.character(BP),
         MF = as.character(MF),
         CC = as.character(CC),
         treatment = factor(treatment, levels = c("FASW", "ABS", "Labrenzia", "Ruegeria", "Vibrio", "multi")),
         timepoint = factor(timepoint, levels = c("1dpi", "21dpi")))

# BP stacked bar
plot_pathway_bp <- deg_lfc_cats %>%
  dplyr::count(treatment, timepoint, direction, BP) %>%
  mutate(n_signed = ifelse(direction == "Down", -n, n)) %>%
  ggplot(aes(x = treatment, y = n_signed, fill = BP)) +
  geom_col() +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  scale_fill_manual(values = bp_colors) +
  facet_wrap(~ timepoint) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "DEGs by Biological Process", y = "Count (up / down)", x = NULL)

# MF stacked bar
plot_pathway_mf <- deg_lfc_cats %>%
  dplyr::count(treatment, timepoint, direction, MF) %>%
  mutate(n_signed = ifelse(direction == "Down", -n, n)) %>%
  ggplot(aes(x = treatment, y = n_signed, fill = MF)) +
  geom_col() +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  scale_fill_manual(values = mf_colors) +
  facet_wrap(~ timepoint) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "DEGs by Molecular Function", y = "Count (up / down)", x = NULL)

# CC stacked bar
plot_pathway_cc <- deg_lfc_cats %>%
  dplyr::count(treatment, timepoint, direction, CC) %>%
  mutate(n_signed = ifelse(direction == "Down", -n, n)) %>%
  ggplot(aes(x = treatment, y = n_signed, fill = CC)) +
  geom_col() +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  scale_fill_manual(values = cc_colors) +
  facet_wrap(~ timepoint) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "DEGs by Cellular Component", y = "Count (up / down)", x = NULL)

plot_pathway_bp / plot_pathway_mf / plot_pathway_cc
```

```{r save-pathway-plots}
ggsave("../results/pathway_composition_BP.pdf", plot_pathway_bp, width = 12, height = 5)
ggsave("../results/pathway_composition_MF.pdf", plot_pathway_mf, width = 12, height = 5)
ggsave("../results/pathway_composition_CC.pdf", plot_pathway_cc, width = 12, height = 5)
```

## 8.11 Log2FC Boxplots by GO Category

Distribution of log2 fold-change values across GO categories, showing the direction and magnitude of expression change per functional group. Each box = all DEGs in that category; fill = treatment; facets = timepoint.

```{r lfc-boxplots-bp, fig.width=14, fig.height=6}
plot_lfc_bp <- deg_lfc_cats %>%
  filter(BP != "Unannotated") %>%
  ggplot(aes(x = BP, y = log2FoldChange, fill = treatment)) +
  geom_boxplot(outlier.size = 0.4, position = position_dodge(0.8), alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~ timepoint) +
  scale_fill_manual(values = palette) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Log2FC by Biological Process Category",
       y = "log2 Fold Change (vs ABS)", x = NULL)
plot_lfc_bp
```

```{r lfc-boxplots-mf, fig.width=14, fig.height=6}
plot_lfc_mf <- deg_lfc_cats %>%
  filter(MF != "Unannotated") %>%
  ggplot(aes(x = MF, y = log2FoldChange, fill = treatment)) +
  geom_boxplot(outlier.size = 0.4, position = position_dodge(0.8), alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~ timepoint) +
  scale_fill_manual(values = palette) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Log2FC by Molecular Function Category",
       y = "log2 Fold Change (vs ABS)", x = NULL)
plot_lfc_mf
```

```{r lfc-boxplots-cc, fig.width=14, fig.height=6}
plot_lfc_cc <- deg_lfc_cats %>%
  filter(CC != "Unannotated") %>%
  ggplot(aes(x = CC, y = log2FoldChange, fill = treatment)) +
  geom_boxplot(outlier.size = 0.4, position = position_dodge(0.8), alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~ timepoint) +
  scale_fill_manual(values = palette) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Log2FC by Cellular Component Category",
       y = "log2 Fold Change (vs ABS)", x = NULL)
plot_lfc_cc
```

```{r save-lfc-boxplots}
ggsave("../results/lfc_boxplot_BP.pdf", plot_lfc_bp, width = 14, height = 6)
ggsave("../results/lfc_boxplot_MF.pdf", plot_lfc_mf, width = 14, height = 6)
ggsave("../results/lfc_boxplot_CC.pdf", plot_lfc_cc, width = 14, height = 6)
```

---

# 9. DESeq2 Analysis — Symbiont (*Symbiodiniaceae*)

Same workflow as Section 8, now applied to algal symbiont genes

```{r deseq-sym-setup}
dds_sym <- DESeqDataSetFromMatrix(
  countData = counts_sym,
  colData = colData,
  design = ~ timepoint + treatment)

cat(nrow(dds_sym), "genes before filtering\n")
keep <- rowSums(counts(dds_sym) >= 10) >= 4
dds_sym <- dds_sym[keep, ]
cat(nrow(dds_sym), "genes after filtering\n")

dds_sym <- DESeq(dds_sym)
```

```{r pca-sym}
vsd_sym <- vst(dds_sym, blind = TRUE)

sym_pca <- plotPCA(vsd_sym, intgroup = c("treatment", "timepoint"), returnData = TRUE)
percent_var <- round(100 * attr(sym_pca, "percentVar"))

p <- ggplot(sym_pca, aes(x = PC1, y = PC2, color = treatment, shape = timepoint)) +
  geom_point(size = 3) +
  scale_color_manual(values = palette) +
  theme_bw() +
  labs(x = paste0("PC1: ", percent_var[1], "% variance"),
       y = paste0("PC2: ", percent_var[2], "% variance"),
       title = "PCA — Symbiodinaceae")
p
ggsave("../results/pca_symbiont_all.pdf", p, width = 7, height = 5)
# Symbiont expression affected by ABS at 1dpi; by 21dpi returns toward control
```

```{r deseq-sym-contrasts}
# 1dpi
dds_sym_1dpi <- dds_sym[, dds_sym$timepoint == "1dpi"]
dds_sym_1dpi$timepoint <- droplevels(dds_sym_1dpi$timepoint)
design(dds_sym_1dpi) <- ~ treatment
dds_sym_1dpi <- DESeq(dds_sym_1dpi)

res_sym_1dpi_control <- results(dds_sym_1dpi, contrast = c("treatment", "FASW", "ABS"))
res_sym_1dpi_ISOA    <- results(dds_sym_1dpi, contrast = c("treatment", "Labrenzia", "ABS"))
res_sym_1dpi_ISOB    <- results(dds_sym_1dpi, contrast = c("treatment", "Ruegeria", "ABS"))
res_sym_1dpi_ISOC    <- results(dds_sym_1dpi, contrast = c("treatment", "Vibrio", "ABS"))
res_sym_1dpi_ISOABC  <- results(dds_sym_1dpi, contrast = c("treatment", "multi", "ABS"))

# 21dpi
dds_sym_21dpi <- dds_sym[, dds_sym$timepoint == "21dpi"]
dds_sym_21dpi$timepoint <- droplevels(dds_sym_21dpi$timepoint)
design(dds_sym_21dpi) <- ~ treatment
dds_sym_21dpi <- DESeq(dds_sym_21dpi)

res_sym_21dpi_control <- results(dds_sym_21dpi, contrast = c("treatment", "FASW", "ABS"))
res_sym_21dpi_ISOA    <- results(dds_sym_21dpi, contrast = c("treatment", "Labrenzia", "ABS"))
res_sym_21dpi_ISOB    <- results(dds_sym_21dpi, contrast = c("treatment", "Ruegeria", "ABS"))
res_sym_21dpi_ISOC    <- results(dds_sym_21dpi, contrast = c("treatment", "Vibrio", "ABS"))
res_sym_21dpi_ISOABC  <- results(dds_sym_21dpi, contrast = c("treatment", "multi", "ABS"))
```

``` {r deg-summary-sym}
deg_summary_sym <- bind_rows(
  summarize_degs(res_sym_1dpi_control) %>% mutate(timepoint = "1dpi",  contrast = "FASW_vs_ABS"),
  summarize_degs(res_sym_1dpi_ISOA)    %>% mutate(timepoint = "1dpi",  contrast = "Labrenzia_vs_ABS"),
  summarize_degs(res_sym_1dpi_ISOB)    %>% mutate(timepoint = "1dpi",  contrast = "Ruegeria_vs_ABS"),
  summarize_degs(res_sym_1dpi_ISOC)    %>% mutate(timepoint = "1dpi",  contrast = "Vibrio_vs_ABS"),
  summarize_degs(res_sym_1dpi_ISOABC)  %>% mutate(timepoint = "1dpi",  contrast = "multi_vs_ABS"),
  summarize_degs(res_sym_21dpi_control) %>% mutate(timepoint = "21dpi", contrast = "FASW_vs_ABS"),
  summarize_degs(res_sym_21dpi_ISOA)    %>% mutate(timepoint = "21dpi", contrast = "Labrenzia_vs_ABS"),
  summarize_degs(res_sym_21dpi_ISOB)    %>% mutate(timepoint = "21dpi", contrast = "Ruegeria_vs_ABS"),
  summarize_degs(res_sym_21dpi_ISOC)    %>% mutate(timepoint = "21dpi", contrast = "Vibrio_vs_ABS"),
  summarize_degs(res_sym_21dpi_ISOABC)  %>% mutate(timepoint = "21dpi", contrast = "multi_vs_ABS")
) %>% select(timepoint, contrast, up, down, total)

deg_summary_sym
# ABS treatment creates an acute gene expression change in the algal symbiont, response fades by 21dpi. Single inoculations have little influence on algal symbiont gene expression but in multi inoculation there is some influence (6 up and 2 down regulated genes)
```

```{r export-degs-sym}
# Export symbiont DEGs with annotations from parsed GTF
export_degs_sym <- function(res, annotation, filename) {
  df <- as.data.frame(res) %>%
    rownames_to_column("gene_id") %>%
    filter(padj < 0.05) %>%
    left_join(annotation, by = "gene_id") %>%
    arrange(padj)
  write.csv(df, file = filename, row.names = FALSE)
  return(df)
}

deg_sym_1dpi_control <- export_degs_sym(res_sym_1dpi_control, annot_sym, "../results/DEG_sym_1dpi_control_vs_ABS.csv")
deg_sym_1dpi_ISOA    <- export_degs_sym(res_sym_1dpi_ISOA,    annot_sym, "../results/DEG_sym_1dpi_ISOA_vs_ABS.csv")
deg_sym_1dpi_ISOB    <- export_degs_sym(res_sym_1dpi_ISOB,    annot_sym, "../results/DEG_sym_1dpi_ISOB_vs_ABS.csv")
deg_sym_1dpi_ISOC    <- export_degs_sym(res_sym_1dpi_ISOC,    annot_sym, "../results/DEG_sym_1dpi_ISOC_vs_ABS.csv")
deg_sym_1dpi_ISOABC  <- export_degs_sym(res_sym_1dpi_ISOABC,  annot_sym, "../results/DEG_sym_1dpi_ISOABC_vs_ABS.csv")

deg_sym_21dpi_control <- export_degs_sym(res_sym_21dpi_control, annot_sym, "../results/DEG_sym_21dpi_control_vs_ABS.csv")
deg_sym_21dpi_ISOA    <- export_degs_sym(res_sym_21dpi_ISOA,    annot_sym, "../results/DEG_sym_21dpi_ISOA_vs_ABS.csv")
deg_sym_21dpi_ISOB    <- export_degs_sym(res_sym_21dpi_ISOB,    annot_sym, "../results/DEG_sym_21dpi_ISOB_vs_ABS.csv")
deg_sym_21dpi_ISOC    <- export_degs_sym(res_sym_21dpi_ISOC,    annot_sym, "../results/DEG_sym_21dpi_ISOC_vs_ABS.csv")
deg_sym_21dpi_ISOABC  <- export_degs_sym(res_sym_21dpi_ISOABC,  annot_sym, "../results/DEG_sym_21dpi_ISOABC_vs_ABS.csv")

saveRDS(dds_sym, "../results/dds_sym.rds")
saveRDS(dds_sym_1dpi, "../results/dds_sym_1dpi.rds")
saveRDS(dds_sym_21dpi, "../results/dds_sym_21dpi.rds")
```

---

# 10. Bacterial Gene Expression (*Labrenzia*, *Ruegeria*, *Vibrio*)

## 10.1 DESeq Feasibility Assessment

```{r bacterial-feasibility}

relevant_treatments <- list(
  lab = c("ABS", "Labrenzia", "multi"),
  rue = c("ABS", "Ruegeria", "multi"),
  vib = c("ABS", "Vibrio", "multi")
)

bacterial_counts <- list(lab = counts_lab, rue = counts_rue, vib = counts_vib)

bact_feasibility <- map_dfr(names(bacterial_counts), function(org) {
  mat <- bacterial_counts[[org]]
  rel_samples <- rownames(colData)[colData$treatment %in% relevant_treatments[[org]]]
  mat_rel <- mat[, rel_samples]
  
  data.frame(
    organism = org,
    relevant_samples = ncol(mat_rel),
    median_library_size = median(colSums(mat_rel)),
    genes_detected = sum(rowSums(mat_rel) > 0),
    genes_passing_filter = sum(rowSums(mat_rel >= 5) >= 2)
  )
})

bact_feasibility
# All three fail feasibility (need >10k median library, >100 genes passing filter)
# Labrenzia: 285 reads, 34 genes; Ruegeria: 108 reads, 45 genes; Vibrio: 47 reads, 159 genes
# Library sizes ~100x too low for DESeq2 dispersion estimation
```

## 10.2 dmdA and dddL Search

Search bacterial annotations for DMSP-cycling genes: `dmdA` (DMSP demethylase) and `dddL` (DMSP lyase). Also performs a broader sulfur/DMSP product search as a fallback. TPM values are plotted across all treatments and timepoints for any hits found.

```{r dmdA-dddL-search, fig.width=12, fig.height=8}
# Targeted search: gene name or product description
sulfur_hits <- annot_bact %>%
  filter(
    grepl("dmdA|dddL", gene_name, ignore.case = TRUE) |
    grepl("dmdA|dddL|DMSP demethylase|DMSP lyase|dimethylsulfoniopropionate lyase|dimethylsulfoniopropionate demethylase",
          product, ignore.case = TRUE)
  )

cat("Targeted dmdA/dddL hits:\n")
print(sulfur_hits %>% select(organism, gene_id, gene_name, product))

# Broader DMSP/sulfur catalog
sulfur_broad <- annot_bact %>%
  filter(grepl("sulfur|DMSP|dimethyl|methanethiol|\\bDMS\\b|thiosulfate|sulfo|methionine sulfoxide",
               product, ignore.case = TRUE)) %>%
  select(organism, gene_id, gene_name, product)

cat("\nBroader sulfur/DMSP gene catalog:\n")
print(sulfur_broad)

# Combine and plot TPM for any hits present in count matrix
sulfur_ids <- unique(c(sulfur_hits$gene_id, sulfur_broad$gene_id))
sulfur_ids <- sulfur_ids[sulfur_ids %in% rownames(tpm_matrix)]

if (length(sulfur_ids) > 0) {
  sulfur_annot <- bind_rows(sulfur_hits, sulfur_broad %>%
    left_join(annot_bact %>% select(gene_id, gene_name, product, organism), by = c("gene_id","gene_name","product","organism"))) %>%
    distinct(gene_id, .keep_all = TRUE) %>%
    mutate(gene_label = case_when(
      !is.na(gene_name) & gene_name != "" ~ paste0(gene_name, " (", organism, ")"),
      TRUE ~ paste0(str_trunc(product, 25), " (", organism, ")")
    ))

  tpm_matrix[sulfur_ids, , drop = FALSE] %>%
    as.data.frame() %>%
    rownames_to_column("gene_id") %>%
    pivot_longer(-gene_id, names_to = "sample", values_to = "TPM") %>%
    left_join(metadata, by = "sample") %>%
    left_join(sulfur_annot %>% select(gene_id, gene_label), by = "gene_id") %>%
    mutate(TPM_plot = pmax(TPM, 0.001)) %>%
    ggplot(aes(x = treatment, y = TPM_plot, color = treatment)) +
    geom_jitter(width = 0.2, size = 2, alpha = 0.8) +
    geom_boxplot(outlier.shape = NA, alpha = 0.2, color = "black") +
    facet_grid(gene_label ~ timepoint, scales = "free_y") +
    scale_color_manual(values = palette) +
    scale_y_log10() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.text.y = element_text(size = 7),
          legend.position = "none") +
    labs(title = "dmdA / dddL and Related Sulfur Cycling Genes: TPM",
         y = "TPM (log10 scale)", x = NULL)
} else {
  cat("No dmdA/dddL or related sulfur cycling genes found in count matrix.\n")
}
```

---

# 11. Save Outputs and Session Info

## Save Publication Figures

```{r save-pub-figures}
# Fig 4A — bicluster heatmaps
pdf("../results/Fig4A_bicluster_1dpi.pdf", width = 12, height = 10)
fig4a_1dpi
dev.off()

pdf("../results/Fig4A_bicluster_21dpi.pdf", width = 12, height = 10)
fig4a_21dpi
dev.off()

# Fig 5A-F — volcano panel
fig5_volcano_panel <- (p_fig5a + p_fig5b) / (p_fig5c + p_fig5d) / (p_fig5e_vib + p_fig5f_vib)
ggsave("../results/Fig5_volcanoes_single_vs_community.pdf",
       fig5_volcano_panel, width = 14, height = 16)
ggsave("../results/Fig5_volcanoes_single_vs_community.png",
       fig5_volcano_panel, width = 14, height = 16, dpi = 300)

# Save annotation tables for downstream use
write.csv(annot_bact, "../results/annotation_bacteria.csv", row.names = FALSE)
write.csv(annot_sym, "../results/annotation_symbiodiniaceae.csv", row.names = FALSE)
```

```{r session-info}
sessionInfo()
```
