---
title: "Transcriptomic Analysis - Host DESeq2"
author: "Maggie Sogin"
date: "2026-07-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,           # Show code in the output
  warning = FALSE,       # Hide warning messages
  message = FALSE,       # Hide package loading messages
  fig.width = 10,        # Default figure width (inches)
  fig.height = 6,        # Default figure height (inches)
  fig.align = "center"   # Center all figures
)
```

# 1. Load Libraries & Define Constants

```{r load-libraries}
library(readxl)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(matrixStats)
library(ggrepel)
library(patchwork)
library(UpSetR)
library(stringr)
library(topGO)
library(vegan)
library(dplyr)
library("RColorBrewer")
library(scales)
```

```{r define-constants}
palette <- c(
  "FASW" = 'grey60',
  "ABS"     = 'black',
  "Labrenzia"    = '#A6CEE3',
  "Ruegeria"    = '#1F78B4',
  "Vibrio"    = '#B2DF8A',
  "multi"  = 'violet')
treatment_levels <- c(
  "control" = "FASW", 
  "ABS" = "ABS", 
  "ISOA" = "Labrenzia", 
  "ISOB" = "Ruegeria", 
  "ISOC" = "Vibrio", 
  "ISOABC" = "multi")

path.out<-"../../output/transcriptomics"
```

# 2. Load All Data Files

```{r}
#load data
load(file.path(path.out,'host_data.RData'))
ls()
```


# 3. DESeq2 Analysis — Host (*Aiptasia*)

Host differential expression analysis using DESeq2. DESeq2 models RNA-seq count data using a negative binomial distribution, accounts for library size differences between samples, and uses shrinkage to produce more reliable fold-change estimates.

The design should include three terms: timepoint, teratment and treatment:timepoint interaction

## 3.1 Create DESeq object, filter data and run LRT test

https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#countmat
As per recommendations in Interactions section: 

```{r specify models and generate DESeq Object}
dds_aip <- DESeqDataSetFromMatrix(
  countData = counts_aip,
  colData = colData,
  design = ~ timepoint*treatment)

cat(nrow(dds_aip), "genes before filtering\n")
keep <- rowSums(counts(dds_aip) >= 10) >= 4
dds_aip <- dds_aip[keep, ]
cat(nrow(dds_aip), "genes after filtering\n")
#removes genes that don't have at least 10 counts in at least 4 samples, this helps to reduced noise in analysis 
```

## 3.2 Run DESeq analysis 
want to look at differences in gene expression at each time-point across treatment groups
```{r}
ds<-DESeq(dds_aip)
plotDispEsts(ds)
# Dispersion plot looks good: trend line decreases smoothly, gene-wise estimates converge (black dots)
```


## 3.3 Exploratory plots

### VST Transformation

```{r vst-aip}
vsd_aip <- vst(dds_aip, blind = TRUE)
```

### PCA — All Samples
```{r pca-aip-all}
pca_data <- plotPCA(vsd_aip, intgroup = c("treatment", "timepoint"), returnData = TRUE)
percent_var <- round(100 * attr(pca_data, "percentVar"))

p <- ggplot(pca_data, aes(x = PC1, y = PC2, color = treatment, shape = timepoint)) +
  geom_point(size = 3) +
  scale_color_manual(values = palette) +
  theme_bw() +
  labs(x = paste0("PC1: ", percent_var[1], "% variance"),
       y = paste0("PC2: ", percent_var[2], "% variance"),
       title = "PCA — Aiptasia")
p
ggsave(file.path(path.out,"pca_host_all.pdf"), p, width = 7, height = 5)
# Timepoint is the dominant source of variation, although controls at 1dpi cluster with all 21dpi, suggesting variation largely due to ABS influence
```
Impacts of the antibiotics is negligble at day 21

```{r PERMANOVA}
# get VST matrix (samples as rows)
vsd_mat <- t(assay(vsd_aip))

# pull metadata
meta <- as.data.frame(colData(vsd_aip))

# run PERMANOVA
perm_res <- adonis2(vsd_mat ~ treatment+timepoint,
                    data = meta,
                    method = "euclidean",
                    permutations = 999,
                    by="terms")
perm_res
```
Significant effect of time but not treatment on expression profiles
Suggests that expression differences are related to ABS exposure and not to microbial inoculations. 

### Sampel to Sample distance matrix
```{r}
sampleDists <- dist(vsd_mat)
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd_aip$timepoint, vsd_aip$treatment, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

### Gene expression maps 
```{r}
timepoint_label="1dpi"

#get mat and transform to z-score
mat<-assay(vsd_aip[, vsd_aip$timepoint == timepoint_label])
mat_z<-t(scale(t(mat)))
mat<-mat[complete.cases(mat_z),]

title<-"gene expression heatmap"

# Column annotation
ann_col <- data.frame(
    Treatment = vsd_aip[, vsd_aip$timepoint == timepoint_label]$treatment,
    row.names = colnames(mat_z))

#row annotations
row_lab <- aip_annotate$protein_name[match(rownames(mat_z), aip_annotate$gene_id)]

p <- pheatmap(mat_z,
    annotation_col = ann_col,
    annotation_colors = list(
    Treatment = palette),
    show_rownames = F,
    show_colnames = F,
    labels_row = row_lab,
    fontsize_row = 5.5,
    cutree_rows = 3,
    cutree_cols = 4,
    main = title,
    silent = TRUE)
grid::grid.newpage(); grid::grid.draw(p$gtable)
```

## 3.4 Contrasts
### 1dpi
```{r 1 dpi contrasts}
tp<-'1dpi'
dds_aip_1dpi <- ds[, ds$timepoint == tp]
dds_aip_1dpi$timepoint <- droplevels(dds_aip_1dpi$timepoint)
design(dds_aip_1dpi) <- ~ treatment
dds_aip_1dpi$treatment <- relevel(dds_aip_1dpi$treatment, ref = "ABS")
dds_aip_1dpi <- DESeq(dds_aip_1dpi)  # re-run after releveling

resultsNames(dds_aip_1dpi)

#contrasts
res_aip_1dpi_control <- results(dds_aip_1dpi, name = "treatment_FASW_vs_ABS")
res_aip_1dpi_ISOA <- results(dds_aip_1dpi, name = "treatment_Labrenzia_vs_ABS")
res_aip_1dpi_ISOB <- results(dds_aip_1dpi, name = "treatment_Ruegeria_vs_ABS")
res_aip_1dpi_ISOC <- results(dds_aip_1dpi, name = "treatment_Vibrio_vs_ABS")
res_aip_1dpi_ISOABC <- results(dds_aip_1dpi, name = "treatment_multi_vs_ABS")

#contrasts with lfcShring
res_aip_1dpi_control_shrunk <- lfcShrink(dds_aip_1dpi, coef = "treatment_FASW_vs_ABS", res = res_aip_1dpi_control,type = "apeglm")

res_aip_1dpi_ISOA_shrunk <- lfcShrink(dds_aip_1dpi, coef = "treatment_Labrenzia_vs_ABS", res = res_aip_1dpi_ISOA,type = "apeglm")

res_aip_1dpi_ISOB_shrunk <- lfcShrink(dds_aip_1dpi, coef = "treatment_Ruegeria_vs_ABS", res = res_aip_1dpi_ISOB,type = "apeglm")

res_aip_1dpi_ISOC_shrunk <- lfcShrink(dds_aip_1dpi, coef = "treatment_Vibrio_vs_ABS", res = res_aip_1dpi_ISOC,type = "apeglm")

res_aip_1dpi_ISOABC_shrunk <- lfcShrink(dds_aip_1dpi, coef = "treatment_multi_vs_ABS", res = res_aip_1dpi_ISOABC,type = "apeglm")
```

### 21 dpi
```{r 21 dpi contasts}
tp<-'21dpi'
dds_aip_21dpi <- ds[, ds$timepoint == tp]
dds_aip_21dpi$timepoint <- droplevels(dds_aip_21dpi$timepoint)
design(dds_aip_21dpi) <- ~ treatment
dds_aip_21dpi$treatment <- relevel(dds_aip_21dpi$treatment, ref = "ABS")
dds_aip_21dpi <- DESeq(dds_aip_21dpi)  # re-run after releveling

resultsNames(dds_aip_21dpi)

#contrasts
res_aip_21dpi_control <- results(dds_aip_21dpi, name = "treatment_FASW_vs_ABS")
res_aip_21dpi_ISOA <- results(dds_aip_21dpi, name = "treatment_Labrenzia_vs_ABS")
res_aip_21dpi_ISOB <- results(dds_aip_21dpi, name = "treatment_Ruegeria_vs_ABS")
res_aip_21dpi_ISOC <- results(dds_aip_21dpi, name = "treatment_Vibrio_vs_ABS")
res_aip_21dpi_ISOABC <- results(dds_aip_21dpi, name = "treatment_multi_vs_ABS")

#contrasts with lfcShring
res_aip_21dpi_control_shrunk <- lfcShrink(dds_aip_21dpi, coef = "treatment_FASW_vs_ABS", res = res_aip_21dpi_control,type = "apeglm")

res_aip_21dpi_ISOA_shrunk <- lfcShrink(dds_aip_21dpi, coef = "treatment_Labrenzia_vs_ABS", res = res_aip_21dpi_ISOA,type = "apeglm")

res_aip_21dpi_ISOB_shrunk <- lfcShrink(dds_aip_21dpi, coef = "treatment_Ruegeria_vs_ABS", res = res_aip_21dpi_ISOB,type = "apeglm")

res_aip_21dpi_ISOC_shrunk <- lfcShrink(dds_aip_21dpi, coef = "treatment_Vibrio_vs_ABS", res = res_aip_21dpi_ISOC,type = "apeglm")

res_aip_21dpi_ISOABC_shrunk <- lfcShrink(dds_aip_21dpi, coef = "treatment_multi_vs_ABS", res = res_aip_21dpi_ISOABC,type = "apeglm")
```

## 3.5 DEGs Tables with annotations
GO IDs are swissprot annotations, also included KEGG IDs

```{r - define function to make table}
make_deg_table <- function(comp_data, comparison, timepoint, annotation_table = aip_annotate) {
  res <- data.frame(comp_data) %>%
    filter(padj < 0.05) %>% #change if needed
    rownames_to_column(var = "gene_id") %>%
    left_join(annotation_table, by = "gene_id") %>%
     mutate(comp = comparison, timepoint=timepoint) #%>%
    # mutate(
    #   'go_ids'=GO_IDs..SwissProt., 
    #   'go_cat'=GO_categories..SwissProt., 
    #   'go_bp'=GO_BP..SwissProt.,
    #   'go_mf'=GO_MF..SwissProt.,
    #   'go_cc'=GO_CC..SwissProt.,
    #   'description'=Hit_description..SwissProt.) %>% 
    # dplyr::select(
    #   gene_id, comp,timepoint, baseMean, log2FoldChange, lfcSE, padj,
    #   protein_name, go_ids, description,go_cat,
    #   go_bp, go_mf, go_cc, KEGG_ID
    # )
  return(res)
}
```

###1 dpi
```{r}
deg_set_fasw_v_abs_1dpi<-make_deg_table(res_aip_1dpi_control_shrunk,'FASW_vs_ABS','1dpi',aip_annotate)
deg_set_lab_v_abs_1dpi<-make_deg_table(res_aip_1dpi_ISOA_shrunk,'Labrenzia_vs_ABS','1dpi',aip_annotate)
deg_set_reu_v_abs_1dpi<-make_deg_table(res_aip_1dpi_ISOB_shrunk,'Reugeria_vs_ABS','1dpi',aip_annotate)
deg_set_vib_v_abs_1dpi<-make_deg_table(res_aip_1dpi_ISOC_shrunk,'Vibrio_vs_ABS','1dpi',aip_annotate)
deg_set_multi_v_abs_1dpi<-make_deg_table(res_aip_1dpi_ISOABC_shrunk,'Multi_vs_ABS','1dpi',aip_annotate)

#deg_set_lab_v_multi_1dpi<-make_deg_table(res_aip_1dpi_ISOA_vs_ISOABC,'Labrenzia_vs_multi','1dpi',aip_annotate)
#deg_set_reu_v_multi_1dpi<-make_deg_table(res_aip_1dpi_ISOB_vs_ISOABC,'Reugeria_vs_multi','1dpi',aip_annotate)
#deg_set_vib_v_multi_1dpi<-make_deg_table(res_aip_1dpi_ISOC_vs_ISOABC,'Vibrio_vs_multi','1dpi',aip_annotate)
```

### 21dpi
```{r}
deg_set_fasw_v_abs_21dpi<-make_deg_table(res_aip_21dpi_control_shrunk,'FASW_vs_ABS','21dpi',aip_annotate)
deg_set_lab_v_abs_21dpi<-make_deg_table(res_aip_21dpi_ISOA_shrunk,'Labrenzia_vs_ABS','21dpi',aip_annotate)
deg_set_reu_v_abs_21dpi<-make_deg_table(res_aip_21dpi_ISOB_shrunk,'Reugeria_vs_ABS','21dpi',aip_annotate)
deg_set_vib_v_abs_21dpi<-make_deg_table(res_aip_21dpi_ISOC_shrunk,'Vibrio_vs_ABS','21dpi',aip_annotate)
deg_set_multi_v_abs_21dpi<-make_deg_table(res_aip_21dpi_ISOABC_shrunk,'Multi_vs_ABS','21dpi',aip_annotate)

#deg_set_lab_v_multi_21dpi<-make_deg_table(res_aip_21dpi_ISOA_vs_ISOABC,'Labrenzia_vs_multi','21dpi',aip_annotate)
#deg_set_reu_v_multi_21dpi<-make_deg_table(res_aip_21dpi_ISOB_vs_ISOABC,'Reugeria_vs_multi','21dpi',aip_annotate)
#deg_set_vib_v_multi_21dpi<-make_deg_table(res_aip_21dpi_ISOC_vs_ISOABC,'Vibrio_vs_multi','21dpi',aip_annotate)
```

```{r combine deg tables}
deg_names <- ls(pattern = "^deg_set")
deg_list <- mget(deg_names) #put dataframes into list 
deg_list <- deg_list[!sapply(deg_list, is.null)]# Drop any NULLs
deg_list <- deg_list[sapply(deg_list, nrow) > 0] # Drop any with 0 rows
combined_deg <- bind_rows(deg_list, .id = "source")
write.csv(combined_deg,file.path(path.out,'DEGs.csv'))
combined_deg
```

```{r summary table}
deg_summary<-combined_deg %>% group_by(comp, timepoint) %>% 
  summarize(nr_DEGs=length(gene_id), up=sum(log2FoldChange>0),down=sum(log2FoldChange<0))
deg_summary %>% arrange(timepoint,comp)
```

```{r}
str(combined_deg %>% filter(!comp=="FASW_vs_ABS"))
degs_of_interest<-combined_deg %>% filter(!comp=="FASW_vs_ABS")
```


## 3.6 Visualize Nr DEGs

```{r}
df<-combined_deg %>% separate(comp, into = c('treatment','base'), sep = "_vs_") %>% 
  mutate(direction=ifelse(log2FoldChange>0,'up','down')) 

ggplot(df, aes(x=treatment, y=log2FoldChange, color=direction)) + 
  #geom_hline(yintercept = 1,linetype='dashed', color='gray')+
  #geom_hline(yintercept = -1,linetype='dashed', color='gray')+
  geom_jitter(height=0,width=0.1,alpha=0.5) + facet_grid(base~timepoint) + 
  theme_bw() + xlab('Treatment vs. ABS') 
```

## 3.7 Heatmap of DEGs

- Plot DEGs from each treatment (Labrenzia, Reugeria, Vibrio and Mult) vs ABS comparison 
- Filter genes with padj < 0.05
- we are not plotting DEGs bewteen ASW and FASW as these genes dominate expression differences in system 

#### 1 dpi
```{r}
timepoint_label<-'1dpi'
vsd<-vsd_aip

sig_genes <- unique(c(
  rownames(subset(as.data.frame(res_aip_1dpi_ISOA_shrunk), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOB_shrunk), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOC_shrunk), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_1dpi_ISOABC_shrunk), padj < 0.05))
))

#get mat and transform to z-score
mat<-assay(vsd[, vsd$timepoint == timepoint_label])[sig_genes, ] 
mat_z<-t(scale(t(mat)))
mat<-mat[complete.cases(mat_z),]

title<-cat(timepoint_label, "union DEGs:", length(sig_genes), "->", nrow(mat_z), "plotted\n")

# Column annotation
ann_col <- data.frame(
    Treatment = vsd[, vsd$timepoint == timepoint_label]$treatment,
    row.names = colnames(mat_z))

#row annotations
row_lab <- aip_annotate$protein_name[match(rownames(mat_z), aip_annotate$gene_id)]

p <- pheatmap(mat_z,
    annotation_col = ann_col,
    annotation_colors = list(
    Treatment = palette),
    show_rownames = TRUE,
    show_colnames = F,
    labels_row = row_lab,
    fontsize_row = 5.5,
    #kmeans_k=3,
    cutree_rows = 3,
    cutree_cols = 4,
    main = paste0("DEGs, Aiptasia ", timepoint_label,
                  " (", nrow(mat_z), " genes), Z-score"),
    silent = TRUE)
grid::grid.newpage(); grid::grid.draw(p$gtable)
```
```{r}
tmp<-aip_annotate %>% filter(gene_id %in% sig_genes)
setdiff(sig_genes,tmp$gene_id)
```

extract genes that are up and down regulated with vibrio cluster 
```{r}
# Extract cluster assignments for each gene
cluster_assignments <- cutree(p$tree_row, k = 3)  # match k to cutree_rows above

# View which cluster each gene belongs to
head(cluster_assignments)

# Pull out genes from a specific cluster (e.g. cluster 2)
cluster1_genes <- names(cluster_assignments[cluster_assignments == 1])
cluster2_genes <- names(cluster_assignments[cluster_assignments == 2])
cluster3_genes <- names(cluster_assignments[cluster_assignments == 3])
```

#### 21 dpi
```{r}
timepoint_label<-'21dpi'
vsd<-vsd_aip

sig_genes <- unique(c(
  rownames(subset(as.data.frame(res_aip_21dpi_ISOA), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOB), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOC), padj < 0.05)),
  rownames(subset(as.data.frame(res_aip_21dpi_ISOABC), padj < 0.05))
))

#get mat and transform to z-score
mat<-assay(vsd[, vsd$timepoint == timepoint_label])[sig_genes, ] 
mat_z<-t(scale(t(mat)))
mat<-mat[complete.cases(mat_z),]

title<-cat(timepoint_label, "union DEGs:", length(sig_genes), "->", nrow(mat_z), "plotted\n")

# Column annotation
ann_col <- data.frame(
    Treatment = vsd[, vsd$timepoint == timepoint_label]$treatment,
    row.names = colnames(mat_z))

#row annotations
row_lab <- aip_annotate$protein_name[match(rownames(mat_z), aip_annotate$gene_id)]

p <- pheatmap(mat_z,
    annotation_col = ann_col,
    annotation_colors = list(
    Treatment = palette),
    show_rownames = TRUE,
    show_colnames = F,
    labels_row = row_lab,
    fontsize_row = 5.5,
    cutree_rows = 2,
    cutree_cols = 2,
    main = paste0("DEGs, Aiptasia ", timepoint_label,
                  " (", nrow(mat_z), " genes), Z-score"),
    silent = TRUE)
grid::grid.newpage(); grid::grid.draw(p$gtable)
```

## 3.8 Gene Expression Patterns

```{r}
deg_df <- degs_of_interest %>% 
          dplyr::select(gene_id,protein_name,Protein_ID,
                        Hit_description..SwissProt.,GO_IDs..SwissProt.,
                        GO_categories..SwissProt.,GO_BP..SwissProt.,
                        KEGG_ID) %>% 
          distinct(gene_id, .keep_all = TRUE)
length(unique(degs_of_interest$gene_id))
length(deg_df$gene_id)


head(deg_df)
write.csv(deg_df,'~/Desktop/tmp.csv')
```

GO metabolic processes defined below:
```{r}
category_keywords<-list(
  "immune system" = c("inflammatory","toll","go:0006954","myd88","immune","cy24b","mfsd6"), #GO term for inflation
  "carbohydrate metabolism" = c(),
  "DNA / RNA metabolic processes"  = c("imdh1","pax3","meox1","klf5"),
  "lipid metabolism" = c("go:0006629","go:0120020", "cholesterol","go:0030301","fat3"),
  "cell cycle" = c('adip'),
  "cell proliferation"  = c('brsk2',"go:0051301","fbln4"),
  "catabolic processes" = c('al1a1','tmps9',"mocos",'cah2','cp3ao','netr'), #double check these genes here 
  "signaling" = c("wnt4a","go:0007166","vwde","epha1", "prkx",
                  'aakg2','torso','par16','agrv1',"ctnl1",
                  'fgr1a','ack1','tm198','opsd1','brs4','PIRL'), #and gene regulation
  "stress response" = c("hbap1", "heat shock",'yl89'), #yl89 resonse to AA starvation
  "transporter activity" = c("go:0003333","go:0055085",'mot3','sup9'), #go terms for transprot
  "toxin" = c("phospholipase a2"),
  "metal ion binding"  = c("frrs1","egtb","cp3A0","obl"),
  "programmed cell death" = c('f162a'),
  "symbiosis"=c("sant"),
  "membrane transport" = c('trpl', 'trpc6')
  )
```

```{r assign categories}
assign_category <- function(text, keyword_list) {
  # catch true blanks, NA, and the "NA | | |" artifact from pasting NA protein_name
  if (is.na(text) || str_trim(text) == "" || str_trim(gsub("\\||NA", "", text)) == "") {
    return("Unknown")
  }
  text <- tolower(text)
  hits <- names(keyword_list)[sapply(keyword_list, function(kw) {
    kw <- kw[kw != "" & !is.na(kw)]
    if (length(kw) == 0) return(FALSE)
    any(str_detect(text, kw))
  })]
  
  if (length(hits) == 0) return("Other/Uncategorized")
  if (length(hits) > 1) return(paste(hits, collapse = "; "))
  hits
}

deg_df_annot <- deg_df %>%
  mutate(search_text = paste(protein_name, Hit_description..SwissProt., GO_categories..SwissProt.,sep = " | "),
         category = sapply(search_text, assign_category, keyword_list = category_keywords))

```
```{r}
#manual adjust some entries 
manual_overrides <- tribble(
  ~gene_id,      ~category,   ~override_reason,
  "aip_g1487",  "signaling",  "TrEMBL annotation Notch signaling pathway",
  "aip_g24879", "catabolic processes", "TrEMBL annotation oxidoreductase activity [GO:0016627]",
  "aip_g24878",  "Unknown",             "WAPA -- bacterial toxin-antitoxin gene (B. subtilis), likely spurious/low-confidence BLAST hit, not trustworthy for host functional classification")
  

# apply overrides on top of automatic keyword-based category
deg_df_annot <- deg_df_annot %>%
  left_join(manual_overrides, by = "gene_id", suffix = c("_auto", "")) %>%
  mutate(
    category_final = coalesce(category, category_auto),  # override wins if present, else keep automatic
    was_overridden = !is.na(category)
  ) %>%
  select(-category) %>%
  rename(category_auto = category_auto, category = category_final)

#summary
table(deg_df_annot$category)
```

```{r}
tmp<-degs_of_interest %>% filter(category == "Other/Uncategorized") 
#write.csv(tmp,'~/Desktop/tmp_uncat_full.csv')
```

###Generate plots

```{r - boxplot option}
#get counts
ds_sf<- estimateSizeFactors(ds)
norm_counts <- counts(ds_sf, normalized = TRUE)

deg_ids <- deg_df_annot %>% pull(gene_id) %>% unique()
norm_sub <- norm_counts[rownames(norm_counts) %in% deg_ids, , drop = FALSE]

#get metadata
sample_meta <- as.data.frame(colData(ds_sf)) %>%
  mutate(sample = rownames(colData(ds_sf))) %>%
  select(sample, treatment, timepoint)

#get means and se for each gene
expr_df <- as.data.frame(norm_sub) %>%
  rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "sample", values_to = "norm_expr") %>%
  left_join(sample_meta, by = "sample") %>% 
  group_by(gene_id, treatment, timepoint) %>% 
  summarize(
    mean_avg = mean(norm_expr),
    se       = sd(norm_expr) / sqrt(n()),
    sample_size = n(),
    .groups = "drop"
  )

# pull ABS mean per gene x timepoint as the baseline to divide by
abs_means <- expr_df %>%
  filter(treatment == "ABS") %>%
  select(gene_id, timepoint, abs_mean = mean_avg)

# compute relative mean + relative SE
expr_rel_df <- expr_df %>%
  left_join(abs_means, by = c("gene_id", "timepoint")) %>%
  mutate(
    rel_mean = mean_avg / abs_mean,
    rel_se   = se / abs_mean
  )

# check for zero/near-zero ABS baselines -- these will produce Inf/NaN
expr_rel_df %>% filter(is.infinite(rel_mean) | is.nan(rel_mean))

# add categories
gene_labels <- deg_df_annot %>%
  select(gene_id, category, protein_name) %>%
  distinct(gene_id, .keep_all = TRUE)

expr_rel_plot <- expr_rel_df %>%
  left_join(gene_labels, by = "gene_id") %>%
  mutate(treatment = factor(treatment, levels = c("ABS", setdiff(unique(treatment), "ABS")))) %>% 
  mutate(
    gene_label  = if_else(is.na(protein_name) | protein_name == "", gene_id, protein_name)
  )

## plot
# one timepoint at a time
tp <- "1dpi"
expr_rel_plot <- expr_rel_plot %>%
  mutate(treatment = factor(treatment,
                             levels = c("ABS", "FASW", "multi", "Labrenzia", "Ruegeria", "Vibrio")))


p_rel_1 <- ggplot(expr_rel_plot %>% filter(timepoint == tp),
                 aes(x = treatment, y = rel_mean, group = gene_id, color = gene_label)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  geom_line(linewidth = 0.8, position = position_dodge(width = 0.3)) +
  geom_errorbar(aes(ymin = rel_mean - rel_se, ymax = rel_mean + rel_se),
                width = 0.15, linewidth = 0.5, position = position_dodge(width = 0.3)) +
  geom_point(size = 2, position = position_dodge(width = 0.3)) +
  facet_wrap(~ category, scales = "free_y", ncol = 4) +
  labs(x = NULL, y = "Expression relative to ABS (mean \u00b1 SE)", color = "Gene",
       title = paste("DEG expression relative to ABS \u2014", tp)) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold", size = 9),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        legend.text = element_text(size = 8))

p_rel_1

```
```{r}
# one timepoint at a time
tp <- "21dpi"

p_rel_21 <- ggplot(expr_rel_plot %>% filter(timepoint == tp),
                 aes(x = treatment, y = rel_mean, group = gene_id, color = gene_label)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  geom_line(linewidth = 0.8, position = position_dodge(width = 0.3)) +
  geom_errorbar(aes(ymin = rel_mean - rel_se, ymax = rel_mean + rel_se),
                width = 0.15, linewidth = 0.5, position = position_dodge(width = 0.3)) +
  geom_point(size = 2, position = position_dodge(width = 0.3)) +
  facet_wrap(~ category, scales = "free_y", ncol = 4) +
  labs(x = NULL, y = "Expression relative to ABS (mean \u00b1 SE)", color = "Gene",
       title = paste("DEG expression relative to ABS \u2014", tp)) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold", size = 9),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        legend.text = element_text(size = 8))

p_rel_21
```


```{r}
p_rel_both_tp <- ggplot(expr_rel_plot,
                          aes(x = treatment, y = rel_mean,
                              group = interaction(gene_id, timepoint),
                              color = gene_label, shape = timepoint)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  geom_line(aes(linetype = timepoint), linewidth = 0.7,
            position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = rel_mean - rel_se, ymax = rel_mean + rel_se),
                width = 0.15, linewidth = 0.4,
                position = position_dodge(width = 0.4)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.4)) +
  facet_wrap(~ category, scales = "free_y", ncol = 4) +
  scale_shape_manual(values = c("1dpi" = 16, "21dpi" = 17)) +   # filled circle vs triangle
  labs(x = NULL, y = "Expression relative to ABS (mean \u00b1 SE)",
       color = "Gene", shape = "Timepoint", linetype = "Timepoint",
       title = "DEG expression relative to ABS \u2014 1dpi vs 21dpi") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold", size = 9),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        legend.text = element_text(size = 8))

p_rel_both_tp
```

*some general conclusions here*
1. Inoculation with specific bacteria depresses host expression of key immune genes, this isn't impacted by the multi species infection and disappears at day 21


```{r session-info}
sessionInfo()
```
